---
title: "Intermediate R Lab using QCRC Data"
format: html
editor: visual
---

# Introduction

In this lab, we’ll analyze the **QCRC_FINAL_Deidentified.xlsx** dataset to practice:

-   Two-sample t-test
-   Chi-square test
-   Linear regression
-   Logistic regression
-   Writing `if` statements
-   Iteration using `for` loops and `apply`

Let’s get started!

------------------------------------------------------------------------

# Part 1 - Loading Data

First, load the required packages and read the Excel file.

```{r}
# Install if not already installed \# install.packages("readxl")

library(readxl)
library(tidyverse)
library(janitor)

# Adjust the path if needed

data <-  read_excel(here::here("data","QCRC_FINAL_Deidentified.xlsx"))|>
	mutate(BMI = as.numeric(BMI),
				 `Vent LOS` = as.numeric(`Vent LOS`)) |>
	clean_names()

# View first few rows

head(data)

# View variable names

names(data)

# Quick summary

summary(data)
```

# Part 2 - Two-Sample t-test

Suppose we want to compare the mean Age of patients who were Intubated vs. those who were not.

```{r}
# Convert Intubated to a factor for clarity
data$Intubated <- factor(data$intubated, labels = c("No", "Yes"))

# Check group means
tapply(data$age, data$intubated, mean, na.rm = TRUE)

# Run t-test
t.test(age ~ intubated, data = data)
```

# Exercise

-   Run the above code.

-   Now insert a new code chunk below and write the code to test whether BMI differs between patients who were Intubated vs. those who were not.

    ```{r}
    t.test(bmi~intubated, data=data)
    ```

# Part 3 - Chi-square Test

Let’s check if `CRRT` use is associated with mortality (`Died`).

```{r}
options(scipen=999)
# Convert to factors
data$crrt <- factor(data$crrt, labels = c("No", "Yes"))
data$died <- factor(data$died, labels = c("No", "Yes"))

# Create a contingency table
tab <- table(data$crrt, data$died)
tab

# Perform chi-square test
chisq.test(tab)
```

## Exercise

-   Run the test above.

-   Now insert a new code chunk below and write the code to repeat the chi-square test for `Intubated` vs. `Died`.

```{r}
chisq.test(data$intubated, data$died)
```

# Part 4 - Linear Regression

We’ll model `ICU_LOS` as a function of `Age`.

```{r}
# Fit linear model
lm_model <- lm(icu_los ~ age, data = data)

# View results
summary(lm_model)
```

## Exercise

-   Run the code above.

-   Now, add a chuck and write the code to fit a model predicting `Vent LOS` using `BMI` adjust the analysis to account for `Age` by adding in an additional covariate.

```{r}
model <- glm(vent_los ~ bmi + age +age*bmi, data=data)

summary(model)
```

# Part 5 - Logistic Regression

Let’s predict the odds of dying (`Died`) based on `Age` and `ICU_LOS`.

```{r}
# Fit logistic regression (notice the family)
log_model <- glm(Died ~ Age + ICU_LOS, data = data, family = binomial)

# View summary
summary(log_model)

# View odss ratios
exp(coef(log_model))

```

## Exercise

-   Run the model above.

-   Now, insert a new chunk ande try predicting `Intubated` using `BMI` and `Age`.

# Part 6 - If Statements

`if` lets you run code when a condition is true.

Example:

```{r}
# Example using the first patient
patient_age <- data$Age[1]

if (patient_age > 45) {
  print("Patient is older than 45.")
} else {
  print("Patient is 45 or younger.")
}
```

You can also use `else if`:

```{r}
# Example with more categories
if (patient_age > 65) {
  print("Patient is elderly.")
} else if (patient_age > 45) {
  print("Patient is middle-aged.")
} else {
  print("Patient is young.")
}
```

## Exercise

-   Write an if-else statement that tests if a patient’s age is greater than 45 and prints:

    -   “Patient is older than 45.” if true

    -   Otherwise prints “Patient is 45 or younger.”

```{r}
age_value <- 72

if (age_value > 65) {
  print("Patient is elderly")
} else {
  print("Patient is not elderly")
}
```

# Great work! You have practiced:

-   Hypothesis testing

-   Regression models

-   Conditional logic
