---
title: "Optional Advanced: Likelihoods with optim"
format:
  revealjs:
    toc: false
    smaller: true
    scrollable: true
---

```{r}
#| label: setup
#| include: false

library(printr)
```

## Why this is optional

This deck preserves the maximum-likelihood material from the earlier combined
ODE/optimization deck. It is useful take-home material for learners who want to
fit custom likelihoods, but it is not part of the Day 3 ODE core.

## Learning goals

1. Write a small objective function.
1. Use `optim()` with parameter bounds.
1. Estimate a mosquito infection rate from pooled surveillance data.
1. Extend the likelihood to a two-parameter mixture model.

## What is `optim()`?

`optim()` is R's general-purpose interface for numerical optimization.

It expects a function to minimize. If you want to maximize a likelihood, it is
common to minimize the negative log likelihood.

```r
optim(
  par = starting_values,
  fn = objective_function,
  method = "L-BFGS-B",
  lower = lower_bounds,
  upper = upper_bounds
)
```

## Warm-up

Find the maximum of:

$$
f(x) = \sin(x)e^{-x^2}
$$

```{r}
curve(
  sin(x) * exp(-x^2),
  from = -2,
  to = 2,
  n = 1000,
  xlab = "x",
  ylab = "f(x)"
)
```

## Warm-up solution

```{r}
warmup_fit <- optim(
  par = 0,
  fn = \(x) sin(x) * exp(-x^2),
  method = "BFGS",
  control = list(fnscale = -1)
)

warmup_fit
```

```{r}
curve(
  sin(x) * exp(-x^2),
  from = -2,
  to = 2,
  n = 1000,
  xlab = "x",
  ylab = "f(x)"
)
abline(v = warmup_fit$par, col = "red", lwd = 2)
abline(h = warmup_fit$value, col = "red", lwd = 2)
```

## Pooled mosquito surveillance

Suppose a health department monitors West Nile virus in mosquito pools.

- 20 pools were tested.
- Each pool contains 10 mosquitoes.
- 7 pools were positive.

If the individual mosquito infection risk is `p`, the probability a pool tests
positive is:

$$
P(+) = 1 - (1 - p)^{10}
$$

## Write the likelihood

For 7 positive pools out of 20:

$$
Y \sim Binomial(20, 1 - (1 - p)^{10})
$$

In R, we can write the negative log likelihood directly.

```{r}
pooled_nll <- function(p) {
  pool_positive_probability <- 1 - (1 - p)^10

  -dbinom(
    x = 7,
    size = 20,
    prob = pool_positive_probability,
    log = TRUE
  )
}
```

## Estimate the infection risk

```{r}
pooled_fit <- optim(
  par = 0.05,
  fn = pooled_nll,
  method = "L-BFGS-B",
  lower = 0.00001,
  upper = 0.99999,
  hessian = TRUE
)

pooled_fit
```

```{r}
infection_risk <- pooled_fit$par
infection_risk
```

## Approximate confidence interval

For a one-parameter likelihood, a quick Wald-style interval uses the Hessian.

```{r}
pooled_se <- sqrt(1 / pooled_fit$hessian[1, 1])
pooled_ci <- infection_risk + c(-1.96, 1.96) * pooled_se

c(
  estimate = infection_risk,
  lower = pooled_ci[1],
  upper = pooled_ci[2]
)
```

This approximation is convenient, but it is not magic. Check whether the
parameter is near a boundary before trusting it.

## Practice checkpoint: zero inflation

**Prompt**

Some trap regions may have no West Nile virus introduced yet. Let `kappa` be
the probability that the region is exposed to the risk process at all.

Use:

$$
P(+) = \kappa \left(1 - (1 - p)^{10}\right)
$$

Estimate `p` and `kappa` from the same 7 positive pools out of 20.

## Practice checkpoint: hint

Your function should take a length-two vector:

```r
zip_nll <- function(par) {
  p <- par[1]
  kappa <- par[2]

  ...
}
```

Use length-two `par`, `lower`, and `upper` values in `optim()`.

## Practice checkpoint: solution

```{r}
zip_nll <- function(par) {
  p <- par[1]
  kappa <- par[2]

  pool_positive_probability <- kappa * (1 - (1 - p)^10)

  -dbinom(
    x = 7,
    size = 20,
    prob = pool_positive_probability,
    log = TRUE
  )
}

zip_fit <- optim(
  par = c(p = 0.05, kappa = 0.70),
  fn = zip_nll,
  method = "L-BFGS-B",
  lower = c(0.00001, 0.00001),
  upper = c(0.99999, 0.99999)
)

zip_fit$par
```

With only one aggregate count, many combinations of `p` and `kappa` can produce
similar fits. The model is easy to code, but the data may not identify both
parameters well.

## Take-home extension

The local file `data/mos-nyc.csv` contains county-level mosquito pool data.

```{r}
mosquito_pools <- read.csv(
  file.path("..", "data", "mos-nyc.csv"),
  check.names = FALSE
)

head(mosquito_pools)
```

**Prompt**

Assume each submitted pool contains 50 mosquitoes. Estimate a single mosquito
infection risk using all counties together.

**Solution outline**

Sum `PCR Positive` and `Pools submitted` across counties, then reuse
`pooled_nll()` with `x`, `size`, and pool size as arguments instead of fixed
numbers.
