---
title: "Module 4: Functions"
format: 
  revealjs:
    scrollable: true
    toc: false
---

## Learning Objectives

After module XX, you should be able to:

- Name the parts of a function
- Write a function
- Use the R/RStudio debugging tools for functions

## Writing your own functions

**why create your own function?**

1. to cut down on repetitive coding
2. to organize code into manageable chunks
3. to avoid running code unintentionally
4. to use names that make sense to you

## Writing your own functions

Here we will write a function that multiplies some number (x) by 2:

```{r eval=TRUE, echo=TRUE}
times_2 <- function(x) x*2
```
When you run the line of code above, you make it ready to use (no output yet!)
Let's test it!

```{r eval=TRUE, echo=TRUE}
times_2(x=10)
```

## Writing your own functions: { }

Adding the curly brackets - `{ }` - allows you to use functions spanning multiple lines:


```{r eval=TRUE, echo=TRUE}
times_3 <- function(x) {
  x*3
}
times_3(x=10)
```

## Writing your own functions: `return`

If we want something specific for the function's output, we use `return()`. Note, if you want to return more than one object, you need to put it into a list using the `list()` function.

```{r eval=TRUE, echo=TRUE}
times_4 <- function(x) {
  output <- x * 4
  return(list(output, x))
}
times_4(x = 10)
```

- R will "implicitly" return the last evaluated expression by default. 
- You should get in the habit of using "explicit" return statements so that you don't get unexpected output.

## Function Syntax

This is a brief introduction. The syntax is:

```
functionName = function(inputs) {
< function body >
return(list(value1, value2))
}
```

Note to create the function for use you need to 

1. Code/type the function
2. Execute/run the lines of code

Only then will the function be available in the Environment pane and ready to use.

## Writing your own functions: multiple arguments

Functions can take multiple arguments / inputs. Here the function has two arguments `x` and `y`

```{r eval=TRUE, echo=TRUE}
times_2_plus_y <- function(x, y) {
  out <- x * 2 + y
  return(out)
}
times_2_plus_y(x = 10, y = 3)
```

## Writing your own functions: arugment defaults

Functions can have default arguments. This lets us use the function without specifying the arguments

```{r eval=TRUE, echo=TRUE}
times_2_plus_y <- function(x = 10, y = 3) {
  out <- x * 2 + y
  return(out)
}
times_2_plus_y()
```

We got an answer b/c we put defaults into the function arguments.

## Writing a simple function

Let's write a function, `sqdif`, that:

1. takes two numbers `x` and `y` with default values of 2 and 3.
2. takes the difference
3. squares this difference
4. then returns the final value

```
functionName = function(inputs) {
< function body >
return(list(value1, value2))
}
```

```{r eval=TRUE, echo=TRUE}
sqdif <- function(x=2,y=3){
     output <- (x-y)^2
     return(output)
}

sqdif()
sqdif(x=10,y=5)
sqdif(10,5)
```

. . .

* Note that this function is implicitly vectorized even if we didn't mean for it to be.

```{r}
sqdif(c(2, 3, 4), c(7, 3.14, 98))
```

## Writing your own functions: characters

Functions can have any kind of data type input. For example, here is a function with characters:

```{r eval=TRUE, echo=TRUE}
loud <- function(word) {
  output <- rep(toupper(word), 5)
  return(output)
}
loud(word = "hooray!")
```

## The `...` (elipsis) argument {.incremental}

* What if we want our function to take in an arbitrary amount of numbers?
* As an example, let's consider a function that calculates the **geometric mean**.

$$
\text{geometric mean}(\mathbf{x}) = \sqrt[n]{x_1 \cdot x_2 \cdot \ldots \cdot x_n} = \exp\left( \frac{1}{n}\left(\log x_1 + \ldots + \log x_n \right) \right)
$$

* We can do this on a vector, of course.

```{r}
geo_mean <- function(x) {
	output <- exp(mean(log(x)))
	return(output)
}

geo_mean(c(2, 3, 4, 5))
```

## The `...` (elipsis) argument {.incremental}

* But sometimes it's easier to write this a different way.

```{r}
geo_mean <- function(...) {
	x <- c(...)
	output <- exp(mean(log(x)))
	return(output)
}

geo_mean(2, 3, 4, 5)
```

* `...` is a placeholder for as many arguments as you want.

## `...` for pass-through {.scrollable}

* The most common use of `...` is to "pass-through" arguments to underlying functions.

* `mean()` has additional arguments, what if we want to use those in our geometric mean calculation?

```{r}
geo_mean <- function(x, trim = 0, na.rm = FALSE) {
		output <- exp(mean(log(x), trim = trim, na.rm = na.rm))
	return(output)
}
```

. . .

Or we can do this an easier way.

## Function debugging tools if you get stuck

- `traceback()`
- `browser()`
- `debug()` and `debugonce()`
- RStudio breakpoints
- **Last resort:** printing and dumping
- More info: [https://adv-r.hadley.nz/debugging.html](https://adv-r.hadley.nz/debugging.html)

## Function instead of Loop {.scrollable}

Now, let's see how we can repeat our loop example from the measles dataset with a function instead.

Here we are going to set up a function that takes our data frame and outputs the median, first and third quartiles, and range of measles cases for a specified country.

```{r eval=TRUE, echo=TRUE}
get_country_stats <- function(df, iso3_code){
	
	country_data <- subset(df, iso3c == iso3_code)
	
	# Get the summary statistics for this country
	country_cases <- country_data$measles_cases
	country_quart <- quantile(
		country_cases, na.rm = TRUE, probs = c(0.25, 0.5, 0.75)
	)
	country_range <- range(country_cases, na.rm = TRUE)
	
	country_name <- unique(country_data$country)
	
	country_summary <- data.frame(
		country = country_name,
		min = country_range[[1]],
		Q1 = country_quart[[1]],
		median = country_quart[[2]],
		Q3 = country_quart[[3]],
		max = country_range[[2]]
	)
	
	return(country_summary)
}
```

Now we can use the function.

```{r eval=TRUE, echo=TRUE}
meas <- readRDS(here::here("data", "measles_final.Rds"))
get_country_stats(df=meas, iso3_code="IND")
get_country_stats(df=meas, iso3_code="PAK")
```


## Summary

- Simple functions take the form:
```
functionName = function(arguments) {
	< function body >
	return(list(outputs))
}
```
- We can specify arguments defaults when you create the function


## You try it!

Create your own function that calculates the incidence rate per $K$ cases in a given country for all years where that country appeared (e.g. incidence per 10,000 people or another number $K$).

Step 1. Determine your arguments.

Step 2. Begin your function by subsetting the data to include only the country specified in the arguments (i.e, `country_data`).

Step 3. Get the years of interest (what errors might you run into here?).

Step 4. Calculate the incidence rate per $K$ people in each year.

Step 5. Return the output neatly.

**Bonus exercise:** allow the country code argument to be optional, or add a new argument that allows the user to only calculate the incidence for certain years.

## Mini Exercise Answer

```{r echo=TRUE}
get_incidence <- function(df, iso3_code, K = 10000){
	
	country_data <- subset(df, iso3c == iso3_code)
	
	incidence <- country_data$measles_cases / country_data$population * K
	names(incidence) <- country_data$year
	
	return(incidence)
}

get_incidence(df=meas, iso3_code="IND")
get_incidence(df=meas, iso3_code="PAK")
```

## Sidenote: combining functions and loops {.scrollable}

- Writing functions is a way to make complex code a lot easier to understand.
- These two loops do exactly the same thing, but the one with the function is much easier to read for most people.

```{r}
my_countries <- c("IND", "PAK", "BGD", "NPL", "BTN")
n <- length(my_countries)
res <- vector(mode = "list", length = n)
```

```{r}
for (i in 1:n) {
	this_country <- my_countries[[i]]
	country_data <- subset(meas, iso3c == this_country)
	incidence <- country_data$measles_cases / country_data$population * 10000
	names(incidence) <- country_data$year
	res[[i]] <- incidence
}
str(res)
```

```{r}
for (i in 1:n) {
	res[[i]] <- get_incidence(meas, my_countries[[i]], 10000)
}
str(res)
```
