---
title: "Module 6: Useful R packages"
subtitle: "gt · gtsummary · sjPlot · table1 · likert · janitor · rio · here · pacman"
author: "Alex Edwards & Zane Billings"
date: today
format: 
  revealjs:
    theme: simple
    slide-number: true
    scrollable: true
    transition: fade
    code-line-numbers: true
    highlight-style: github
    incremental: false
    toc: false
execute:
  eval: false
  echo: true
editor:
  markdown:
    wrap: 72
---

# Welcome

## Agenda {.smaller}

A one-hour tour of the R packages that take you from **messy data** to
**publication-ready tables and reports** — demonstrated on a real,
de-identified ICU dataset.

## Agenda {.smaller}

| Time  | Topic                                     |
|-------|-------------------------------------------|
| 2 min | Why these packages?                       |
| 2 min | The dataset: QCRC ICU cohort              |
| 3 min | `pacman::p_load()` — loading packages     |
| 3 min | `here` — project-relative file paths      |
| 6 min | `rio` — import/export anything            |
| 6 min | `janitor::clean_names()` — tidy names     |
| 7 min | `table1` — "Table 1"                      |
| 6 min | `gt` — grammar of tables                  |
| 8 min | `gtsummary` — summary & regression tables |
| 8 min | `sjPlot` — `tab_model()` & `tab_xtab()`   |
| 6 min | `likert` — ordinal clinical scores        |
| 3 min | Wrap-up & resources                       |

## Why these packages?

-   Every analysis ends with a table, a report, or a slide someone else
    has to read
-   Base R output (`summary()`, `str()`) is fast to write but ugly to
    share
-   These packages sit **on top of** your normal analysis and turn it
    into something you can paste into a paper, a report, or an email

## Which packages?

::: incremental
-   `here`, `rio`, `janitor` → find, load, and clean your data
-   `table1` → describe your sample
-   `gt`, `gtsummary`, `sjPlot` → build the tables people actually read
-   `likert` → visualize ordered/scale data
-   `pacman` → glue that loads everything with one line
:::

------------------------------------------------------------------------

## About the Dataset

Every example in this deck uses the same real, de-identified file:
**`QCRC_FINAL_Deidentified.xlsx`**

-   288 patients admitted to the ICU with COVID-19
-   10 sheets — `Main_Dataset` (demographics, treatments, outcomes),
    daily `SOFA` organ-failure scores, `Long_Labs`, `Long_Vitals`,
    ventilator settings, medications, and more
-   We'll mainly use `Main_Dataset`: 37 variables including age, sex,
    race, BMI, intubation, CRRT/dialysis, and mortality
-   98 of 288 patients (34%) died — a natural grouping variable for
    every table in this lecture

::: notes
Before running this deck live: save the workbook as
`data/QCRC_FINAL_Deidentified.xlsx` inside your project folder — the
`here()` and `rio::import()` calls later assume that path.
:::

# Setup: `pacman::p_load()`

## The problem with `library()`

A typical analysis script starts like this:

```{r}
library(here)
library(rio)
library(janitor)
library(table1)
library(gt)
library(gtsummary)
library(sjPlot)
library(likert)
library(tidyverse)
```

If **one** of these isn't installed, the whole script stops with an
error — and you have to go install it manually, then re-run.

## `pacman::p_load()`

`pacman` is a package-management package. Its star function,
**`p_load()`**, installs anything missing, then loads everything — in
one line.

```{r}
# install.packages("pacman")   
# only needed once

pacman::p_load(
  here, rio, janitor, table1,
  gt, gtsummary, sjPlot, likert,
  tidyverse
)
```

## `pacman::p_load()`

-   Checks each package name
-   Installs it from CRAN **only if missing**
-   Loads (`library()`s) all of them
-   Great for shared scripts, teaching, and reproducibility setups

# `here`: Project-Relative Paths

## The problem with `setwd()`

```{r}
# fragile — breaks the moment this script runs on
# a different machine, or is Source()'d vs rendered
setwd("~/Downloads/SISMID2026")
qcrc <- import("data/QCRC_FINAL_Deidentified.xlsx")
```

## `here::here()`

```{r}
library(here)

#This is where it think we are
here()

#This gets us to our data
here("data", "QCRC_FINAL_Deidentified.xlsx")

#How do we read in data
qcrcHere <- import(here("data", "QCRC_FINAL_Deidentified.xlsx"))
```

## `here::here()`

-   Finds your project root automatically (looks for a `.Rproj` file,
    `.git` folder, or similar) and builds paths **from there**
-   The same `here()` call works identically whether you run it from
    RStudio, the terminal, or a rendered Quarto document
-   Pairs perfectly with `rio::import()` — see next section

::: notes
Rule of thumb: replace every `setwd()` and every hard-coded absolute
path with `here()`.
:::

# `rio`: Import & Export Anything

## One function in, one function out

`rio` wraps dozens of format-specific packages (`haven`, `readxl`,
`data.table`, `jsonlite`, ...) behind two simple verbs.

```{r}
#| eval: true
library(rio)
library(here)

qcrc <- import(
  here("data", "QCRC_FINAL_Deidentified.xlsx"),
  sheet = "Main_Dataset"
)

dim(qcrc)
```

## One function in, one function out

No more remembering `read_csv()` vs `read.spss()` vs `read_excel()` —
and no hard-coded paths, thanks to `here()`.

## Exporting is just as easy

```{r}
export(qcrc, here("data", "qcrc_clean.xlsx"))
export(qcrc, here("data", "qcrc_clean.rds"))
export(qcrc, here("data", "qcrc_clean.csv"))
```

## Exporting is just as easy

-   Format is inferred from the file extension
-   Supports 30+ formats: `.csv`, `.xlsx`, `.sav`, `.dta`, `.sas7bdat`,
    `.json`, `.rds`, `.parquet`, and more
-   `import_list()` reads **every sheet** at once — perfect for this
    workbook, which has 10 sheets:

```{r}
#| eval: true
#| warning: false
#| message: false

sheets <- import_list(here("data", "QCRC_FINAL_Deidentified.xlsx"))
names(sheets)
```

::: notes
Good moment for a live demo — scroll through `sheets` to show the whole
workbook loaded in one call.
:::

# `janitor`: Clean Names Fast

## Messy names are the #1 headache

```{r}
#| eval: true
names(qcrc)
```

## Messy names are the #1 headache

Notice the inconsistency: `Title Case With Spaces` (`Vent LOS`,
`Peak troponin`), `ALL_CAPS_WITH_UNDERSCORES` (`ICU_ADMIT_DEID`), and
even a colon (`P:F ratio at ICU admission`). All are painful to type or
reference in code.

## `clean_names()`

```{r}
#| eval: true
library(janitor)

qcrc_clean <- qcrc |> clean_names()
names(qcrc_clean)
```

## `clean_names()`

-   Converts everything to consistent `snake_case`
-   Strips spaces and punctuation (even the colon), collapses repeated
    separators, and prefixes names that start with a digit (`30D_...` →
    `x30d_...`)
-   One line, run right after `import()` — every variable name used for
    the rest of this lecture assumes this has already happened

## Bonus: fixing the `"."` missing code {.smaller}

Like a lot of clinical exports, this dataset stores missing values as a
literal `"."` string — which leaves numeric columns like `bmi`, `wbc`,
and `crp` stuck as text.

```{r}
#| eval: true
library(dplyr)

num_cols <- c("bmi", "wbc", "d_dimer", "crp", "peak_troponin",
              "calculated_cstat")

qcrc_clean <- qcrc_clean |>
  mutate(across(all_of(num_cols),
                ~ na_if(as.character(.x), ".")|> 
  								as.numeric())) |>
  mutate(died = factor(died, labels = c("Survived", "Died")))
```

## Bonus: fixing the `"."` missing code {.smaller}

Not a janitor function, but the natural next step after `clean_names()`
— tidy the *names*, then tidy the *values*. This is the `qcrc_clean`
object every later slide builds on.

# `table1`: Describing Your Sample

## The "Table 1" problem

Nearly every clinical paper starts with a baseline characteristics table
split by outcome or treatment group — this is almost always called
**"Table 1."** Our cohort: 288 ICU patients, 98 of whom (34%) died.

Building it by hand means computing means, SDs, counts, percentages, and
p-values separately for every variable.

## `table1()`

```{r}
#| eval: true
library(table1)

#Table One No Strata
table1(~age+bmi+icu_los+intubated+crrt, data=qcrc_clean)
```

## `table1()`

-   Older age, lower BMI, mechanical ventilation, and CRRT (dialysis)
    were all significantly associated with death in this cohort
-   Printed table can be coerced to a data frame and then piped into
    `gt` for a polished, exportable version.

## `table1() with strata`

```{r}
#| eval: true
#Table one with strata
table1(~age+bmi+icu_los+intubated+crrt|female, data=qcrc_clean)
```

## `table1() with two strata`

```{r}
#| eval: true
#Table One with two strata
table1(~age+bmi+icu_los+intubated+crrt|female*died, data=qcrc_clean)
```

# `gt`: A Grammar of Tables

## Any data frame → a polished table

```{r}
#| eval: true
library(gt)

qcrc_clean |>
  select(patient_deid, age, female, race, bmi, icu_los, died) |>
  head(6) |>
  gt() |>
  tab_header(
    title    = "QCRC ICU Cohort",
    subtitle = "First 6 patients"
  ) |>
  fmt_number(columns = c(bmi, icu_los), decimals = 1) |>
  cols_label(patient_deid = "Patient ID", icu_los = "ICU LOS (days)")
```

## Any data frame → a polished table

-   `gt()` turns a data frame into a table object
-   Everything after is a **layer**: header, formatting, labels,
    footnotes, spanners, styling — piped together

## Why `gt` matters

-   It's the **rendering engine** many other packages (including
    `gtsummary`) build on top of
-   Output looks the same in HTML, PDF, Word, and RTF
-   Useful layers to know:
    -   `tab_spanner()` — group columns under a shared header
    -   `tab_source_note()` — add a footnote/citation
    -   `gtsave(tbl, "table.png")` — export to image, HTML, RTF, or Word

# `gtsummary`: Summary & Regression Tables

## `tbl_summary()` — descriptive tables

```{r}
#| eval: true
library(gtsummary)

qcrc_clean |>
  select(age, bmi, wbc, crp, died) |>
  tbl_summary(
    by        = died,
    statistic = list(all_continuous() ~ "{median} ({p25}, {p75})")
  ) |>
  add_p() |>
  add_overall()
```

## `tbl_summary()` — descriptive tables

-   Median (IQR) instead of mean (SD) — a better default for skewed labs
    like CRP and WBC
-   Same goal as `table1`, but the output is already a polished **gt
    table** — ready to drop into a report
-   `add_p()` adds group-comparison p-values; `add_overall()` adds a
    combined column

## `tbl_regression()` — model tables

```{r}
#| eval: true
mod <- glm(died ~ age + bmi + intubated + crrt,
           data = qcrc_clean, family = binomial)

tbl_regression(mod, exponentiate = TRUE) |>
  add_glance_source_note()
```

## `tbl_regression()` — model tables

-   Turns any `lm`/`glm`/`coxph`/mixed model into a clean estimate table
    (odds ratio, CI, p-value) — `exponentiate = TRUE` reports odds
    ratios instead of log-odds
-   `add_glance_source_note()` appends model fit stats (AIC, n)
-   `tbl_merge()` and `tbl_stack()` combine multiple models/tables side
    by side or on top of each other — handy for multi-model comparisons

::: notes
gtsummary is the "batteries included" option — sensible defaults,
statistical tests built in, minimal code. This is the same
`died ~ age + bmi + intubated + crrt` model sjPlot will show next, just
formatted differently.
:::

# `sjPlot`: Regression & Crosstab Tables

## `tab_model()` — regression output

```{r}
#| eval: true
library(sjPlot)

tab_model(
  mod,
  show.ci  = TRUE,
  transform = "exp",
  dv.labels = "Died (ICU, 0/1)"
)
```

## `tab_model()` — regression output

-   Formats coefficients, CIs, and p-values automatically — here as odds
    ratios (`transform = "exp"`) since this is a logistic model
-   Works with `lm`, `glm`, `lmer`, `coxph`, and more
-   Pass **multiple models** as extra arguments to compare them side by
    side in one table

## `tab_xtab()` — cross-tabulations

```{r}
#| eval: true
tab_xtab(
  var.row      = qcrc_clean$race,
  var.col      = qcrc_clean$died,
  show.row.prc = TRUE,
  show.summary = TRUE
)
```

-   Builds a two-way frequency table for two categorical variables
-   `show.row.prc` / `show.col.prc` add row/column percentages
-   `show.summary` appends the chi-square test statistic and p-value
-   Same polished HTML output style as `tab_model()`

::: notes
If time is short, this is the pair most useful to clinical audiences:
one function for regression, one for crosstabs, both publication-ready
with zero manual formatting.
:::

# `likert`: Ordinal / Scale Data

## Beyond survey questions

The `likert` package is built for survey agreement scales, but it works
for **any** ordered categorical variable measured on a shared scale. The
`SOFA` sheet has six organ-system subscores, each rated 0 (normal) to 4
(severe failure) — structurally identical to a 5-point Likert item.

```{r}
#| eval: true
sofa <- import(
  here("data", "QCRC_FINAL_Deidentified.xlsx"),
  sheet = "SOFA"
) |>
  clean_names()

lvls <- as.character(0:4)

sofa_items <- sofa |>
  transmute(
    Coagulation    = factor(coag,  levels = lvls),
    Liver          = factor(liver, levels = lvls),
    CNS            = factor(cns,   levels = lvls),
    Renal          = factor(renal, levels = lvls),
    Cardiovascular = factor(cvs,   levels = lvls),
    Respiratory    = factor(resp,  levels = lvls)
  )
```

## `likert()` + `plot()`

```{r}
#| eval: true
library(likert)

l <- likert(as.data.frame(sofa_items))
summary(l)
```

```{r}
#| eval: true
#| fig-height: 3.5
plot(l)
```

-   Renal and cardiovascular subscores show the most severe dysfunction
    in this cohort (large "3"/"4" categories)
-   `summary()` gives the % of patients per score, per organ system;
    `plot()` produces the classic diverging stacked bar chart
-   The exact same workflow applies to real Likert survey items —
    agreement scales are just another ordered factor with a shared level
    set

# Putting It Together

## Which package for which job? {.smaller}

| Package     | Best for                                                   |
|-------------|------------------------------------------------------------|
| `pacman`    | Loading/installing packages in one line                    |
| `here`      | Project-relative file paths, no more `setwd()`             |
| `rio`       | Reading/writing any data file format                       |
| `janitor`   | Cleaning column names & quick data checks                  |
| `table1`    | Descriptive "Table 1" split by group                       |
| `gt`        | Full control over any custom table's look                  |
| `gtsummary` | Fast, statistically-aware summary & model tables           |
| `sjPlot`    | Regression tables (`tab_model`) and crosstabs (`tab_xtab`) |
| `likert`    | Visualizing ordered survey/scale data                      |

## A typical workflow

```{r}
pacman::p_load(here, rio, janitor, table1, gt, gtsummary, sjPlot, likert)

qcrc <- import(here("data", "QCRC_FINAL_Deidentified.xlsx"),
                sheet = "Main_Dataset") |>
  clean_names()

table1(~age+bmi+female+icu_los+intubated+crrt|died, data=qcrc_clean)
qcrc |>
  tbl_summary(by = died) 

mod <- glm(died ~ age + intubated + crrt,
           data = qcrc, family = binomial)
tab_model(mod)
```

Six packages, six lines, from a raw ICU spreadsheet to publication-ready
output.

# Wrap-Up

## Resources

-   `here`: <https://here.r-lib.org>
-   `rio`: <https://gesistsa.github.io/rio>
-   `janitor`: <https://sfirke.github.io/janitor>
-   `table1`: CRAN vignette —
    `vignette("introduction", package = "table1")`
-   `gt`: <https://gt.rstudio.com>
-   `gtsummary`: <https://www.danieldsjoberg.com/gtsummary>
-   `sjPlot`: <https://strengejacke.github.io/sjPlot>
-   `likert`: CRAN vignette — `vignette("likert", package = "likert")`
-   `pacman`: <https://github.com/trinker/pacman>

## Questions?
