---
title: "Module 5: FunctionalProgramming"
format:
  revealjs:
    toc: false
---

```{r}
#| label: setup
#| include: false

library(printr)
```


## Learning goals

1. Define functional programming
1. Use functional programming tools in base R to repeat multiple functional calls.
1. Use the `purrr` package for easily mapping and reducing functions.

## What is functional programming?

. . .

- All of our code is functional if it does its job?

. . .

- Yes, but we are using the technical definition of functional programming: programming that is based on functions.

. . .

- Specifically, we want to be able to compose functions, which you might remember hating in precalculus.

$$g \circ f = g(f(x))$$

- In computer science, a **functional** is a function that accepts a function as an argument.

## Functional vs. imperative programming

- So far, we've been using **imperative** programming: we save variables, and we update them with new commands. Our code is structured as a list of instructions.

- In **functional** programming, we write everything as a function, and we get the results we want by composing many functions.

. . .

- But we're just going to show you some useful parts of functional programming that you can include in your regular code.

## Ok, but why?

- Recall our function from the previous Module.

```{r eval=TRUE, echo=TRUE}
get_country_stats <- function(df, iso3_code){
	
	country_data <- subset(df, iso3c == iso3_code)
	
	# Get the summary statistics for this country
	country_cases <- country_data$measles_cases
	country_quart <- quantile(
		country_cases, na.rm = TRUE, probs = c(0.25, 0.5, 0.75)
	)
	country_range <- range(country_cases, na.rm = TRUE)
	
	country_name <- unique(country_data$country)
	
	country_summary <- data.frame(
		country = country_name,
		min = country_range[[1]],
		Q1 = country_quart[[1]],
		median = country_quart[[2]],
		Q3 = country_quart[[3]],
		max = country_range[[2]]
	)
	
	return(country_summary)
}
```

. . .

- We could write a loop to get stats for many countries.

```{r eval=TRUE, echo=TRUE}
meas <- readRDS(here::here("data", "measles_final.Rds"))
country_codes <- c("IND", "PAK", "BGD", "NPL")

# Loop setup
out <- vector(mode = "list", length = length(country_codes))
for (i in 1:length(out)) {
	out[[i]] <- get_country_stats(meas, country_codes[[i]])	
}

out
```

. . .

- But if we use a functional programming tool called `lapply`, look how easy it is!

```{r}
out2 <- lapply(country_codes, function(c) get_country_stats(meas, c))
out2
```

- Functional programming techniques can help us avoid writing messy loops and clean up our code.

## Ok, but why?

- R is a functional programming language at its core.
- In R, functions are objects like everything else.
- You never *have* to use FP, but it can help you write neater code.
- More information: https://adv-r.hadley.nz/fp.html.

## Anonymous functions

- In this code, you can see the keyword `function`.

```{r}
out2 <- lapply(country_codes, function(c) get_country_stats(meas, c))
```

. . .

- The function `function(c) get_country_stats(meas, c)` is called an **anonymous function** (or sometimes a lambda).
- It's anonymous because it has no name assigned -- this is just a shortcut for when you only need a function once.
- In R, you can also write `\(c) ...` as a shortcut for writing `function(c) ...` as an anonymous function.

. . .

- This code works just as well, but can be a bit of a hassle.

```{r}
get_country_stats_from_meas <- function(c) get_country_stats(meas, c)
out3 <- lapply(country_codes, get_country_stats_from_meas)
```

## `*apply()` and friends {.incremental}

- R includes many functional programming tools collectively called `*apply()`. Here are the specific ones that are most useful:
- `lapply()`: repeat the function on a list of things.
- `sapply()`: same as `lapply()` but try to simplify the output to a matrix or vector.
- `apply()`: repeat the function over the rows (`margin = 1`) or columns (`margin = 2`) of a matrix.
- `tapply()`: repeat the function over combinations of grouping factors.

## A `tapply()` example {.scrollable}

- `tapply()`: repeat the function over combinations of grouping factors.
- We want to get the average vaccine coverage for each country in the measles dataset -- we need to separate this by the two vaccines as well.
- This is pretty easy to do with `tapply()`.

. . .

```{r}
meas <- readRDS(here::here("data", "measles_final.Rds"))
meas_long <- meas |>
	tidyr::pivot_longer(
		dplyr::starts_with("MCV"),
		names_to = "vaccine_antigen",
		values_to = "vaccine_coverage"
	)
out <- tapply(
	meas_long$vaccine_coverage,
	list(meas_long$iso3c, meas_long$vaccine_antigen),
	FUN = mean
)
head(out)
```

. . .

- Uh oh.

. . .

- If we want to handle the missing values, we can use an anonymous function.

```{r}
out <- tapply(
	meas_long$vaccine_coverage,
	list(meas_long$iso3c, meas_long$vaccine_antigen),
	FUN = \(x) mean(x, na.rm = TRUE)
)
head(out)
```

## You try it with `lapply()`!

- Read in all sheets of the `QCRC_FINAL_Deidentified.xlsx` data using a vector of sheet names and `lapply`.

. . .

- Hint:

```{r, eval = FALSE}
data_file <- here::here("data", "QCRC_FINAL_Deidentified.xlsx")
sheet_names <- readxl::excel_sheets(data_file)
sheet_list <- lapply(
	sheet_names,
	\(name) ...
)
```

. . .

```{r}
data_file <- here::here("data", "QCRC_FINAL_Deidentified.xlsx")
sheet_names <- readxl::excel_sheets(data_file)
sheet_list <- lapply(
	sheet_names,
	\(name) readxl::read_excel(data_file, sheet = name)
)
str(sheet_list, 1)
```



## An `apply()` example

- You'll use `apply()` less often because it's specifically for matrices, where all data types have to be the same.
- First, make a matrix from the QCRC main dataset that only has the columns `Died`, `30D_Mortality`, and `60D_Mortality`.

```{r}
d_matrix <- sheet_list[[1]] |>
	dplyr::select(Died, `30D_Mortality`, `60D_Mortality`) |>
	as.matrix()

head(d_matrix)
```

- Next use `apply` with the correct choice of `MARGIN` and `FUN` to get the proportion of cases that resulted in death at all three time points in one function call.

. . .

- Hint:

```{r, eval = FALSE}
apply(
	d_matrix,
	MARGIN = (1 or 2),
	FUN = some_function
)
```

- Another hint: you can tell if you chose the correct `MARGIN` argument by counting the number of elements in the output. There should be 3, one for each column.

. . .

- Solution:

```{r}
apply(
	d_matrix,
	MARGIN = 2,
	FUN = mean
)
```

## You try it!

- Load the measles dataset. Filter the dataset so you only have records from 2005, and only have complete cases.
- Then, use `split` to split the dataframe into a list by `region`.
- Now, use `lapply()` to fit a Poisson `glm` for each region separately that includes effects of MCV1 and MCV2 coverage.
- Find a way, using `lapply()` or `sapply()`, `coef()`, and `do.call()` to make a nice matrix of the coefficients.

. . .

- Solution for data processing

```{r}
meas <- readr::read_rds(here::here("data", "measles_final.Rds"))
meas_2005 <- meas |>
	dplyr::filter(year == 2005) |>
	tidyr::drop_na()

meas_regions <- split(meas_2005, meas_2005$region)
```

. . .

- Hint

```{r, eval = FALSE}
model_list <- lapply(
	meas_regions,
	\(d) glm(...)
)
model_coefs_list <- lapply(model_list, ...)
model_coefs_mat <- do.call(...)
```


. . .

- Solution

```{r}
model_list <- lapply(
	meas_regions,
	\(d) glm(
		measles_cases ~ MCV1_coverage + MCV2_coverage,
		data = d, family = "poisson"
	)
)
model_coefs_list <- lapply(model_list, coef)
model_coefs_mat <- do.call(rbind, model_coefs_list)

model_coefs_mat
```

## Summary

- Functional programming tools like `*apply()` take a function as an input and use the same function multiple times.
- In R, functions are objects like everything else and manipulating them like objects can help us write readable, fast code.

. . .

- Side note: `purrr` is a modern version of `*apply()` with a more consistent interface. It's worth learning and covered in the [Advanced R book](https://adv-r.hadley.nz/fp.html), but most people still use `*apply()`.
- This book also covers more standard FP tools like reduction and filtering that we didn't have time to talk about.
