---
title: "Optional Advanced: Fitting ODE Models"
format:
  revealjs:
    toc: false
    smaller: true
    scrollable: true
---

```{r}
#| label: setup
#| include: false

library(deSolve)
library(printr)
```

## Why this is optional

This deck parks the old ODE-fitting extension from the combined ODE/optimization
material. It is useful for learners who want to see how simulation code can be
wrapped in an objective function, but it is not part of the Day 3 ODE core.

The point here is the programming pattern, not a recommended inference workflow.

## Data

The observed epidemic curve is local to this repository.

```{r}
observed_sir <- read.csv(file.path("..", "data", "SIR-observed.csv"))
head(observed_sir)
```

```{r}
plot(
  observed_sir$day,
  observed_sir$total_cases,
  pch = 16,
  col = "gray30",
  xlab = "Day",
  ylab = "Observed infected",
  main = "Observed epidemic curve"
)
```

## Reuse the SIR model

```{r}
sir_ode <- function(t, state, parameters) {
  S <- state["S"]
  I <- state["I"]
  R <- state["R"]

  beta <- parameters["beta"]
  gamma <- parameters["gamma"]

  dS <- -beta * S * I
  dI <- beta * S * I - gamma * I
  dR <- gamma * I

  list(c(S = dS, I = dI, R = dR))
}
```

## Write a simulation helper

```{r}
simulate_sir <- function(beta, gamma, times) {
  deSolve::ode(
    y = c(S = 999, I = 1, R = 0),
    times = times,
    func = sir_ode,
    parms = c(beta = beta, gamma = gamma)
  )
}

example_simulation <- simulate_sir(
  beta = 0.001,
  gamma = 0.05,
  times = observed_sir$day
)

head(example_simulation)
```

## Compare one simulation to the data

```{r}
plot(
  observed_sir$day,
  observed_sir$total_cases,
  pch = 16,
  col = "gray30",
  xlab = "Day",
  ylab = "Infected",
  main = "Observed and simulated infected counts"
)
lines(example_simulation[, "time"], example_simulation[, "I"], col = "darkorange2", lwd = 2)
legend(
  "topright",
  legend = c("Observed", "Simulated"),
  col = c("gray30", "darkorange2"),
  pch = c(16, NA),
  lty = c(NA, 1),
  lwd = c(NA, 2),
  bty = "n"
)
```

## Objective function

We need one number that says how badly a parameter pair fits.

Here we use the sum of squared residuals between observed infected counts and
the simulated infected compartment:

$$
SSR = \sum_i \left(I_i - \hat{I}_i(\beta, \gamma)\right)^2
$$

```{r}
sir_ssr <- function(par) {
  beta <- par[1]
  gamma <- par[2]

  if (beta <= 0 || gamma <= 0) {
    return(1e12)
  }

  simulated <- tryCatch(
    simulate_sir(
      beta = beta,
      gamma = gamma,
      times = observed_sir$day
    ),
    error = function(e) NULL
  )

  if (is.null(simulated)) {
    return(1e12)
  }

  predicted_infected <- simulated[, "I"]

  if (any(!is.finite(predicted_infected))) {
    return(1e12)
  }

  fit <- sum((observed_sir$total_cases - predicted_infected)^2)

  if (!is.finite(fit)) {
    return(1e12)
  }

  fit
}
```

## Fit the parameters

```{r}
sir_fit <- optim(
  par = c(beta = 0.001, gamma = 0.05),
  fn = sir_ssr,
  method = "L-BFGS-B",
  lower = c(beta = 0.000001, gamma = 0.000001),
  upper = c(beta = 0.01, gamma = 1)
)

sir_fit$par
```

## Plot the fitted trajectory

```{r}
fitted_simulation <- simulate_sir(
  beta = sir_fit$par["beta"],
  gamma = sir_fit$par["gamma"],
  times = observed_sir$day
)

plot(
  observed_sir$day,
  observed_sir$total_cases,
  pch = 16,
  col = "gray30",
  xlab = "Day",
  ylab = "Infected",
  main = "Observed and fitted SIR trajectory"
)
lines(fitted_simulation[, "time"], fitted_simulation[, "I"], col = "darkorange2", lwd = 2)
legend(
  "topright",
  legend = c("Observed", "Fitted SIR"),
  col = c("gray30", "darkorange2"),
  pch = c(16, NA),
  lty = c(NA, 1),
  lwd = c(NA, 2),
  bty = "n"
)
```

## Practice checkpoint

**Prompt**

Change the starting values in `optim()` to `beta = 0.005` and `gamma = 0.50`.
Do the fitted parameters return to approximately the same values?

**Hint**

Only change the `par = ...` line. Keep the same bounds.

**Solution outline**

Run the same call with:

```r
par = c(beta = 0.005, gamma = 0.50)
```

For this small example, the bounds and data are informative enough that the
result should be very similar. For more complex ODE systems, starting values can
matter a lot.

## Take-home cautions

- Real case data are noisy, delayed, underreported, and often only partly
  observed.
- The observation model matters as much as the ODE model.
- Parameter uncertainty from this quick example is not reliable.
- Use this deck to learn how to connect simulation code to a fitting workflow,
  then reach for a more serious modeling framework for real inference.
