---
title: "Data science walkthrough 2: Repeating an analysis"
format:
  revealjs:
    toc: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## The same problem, repeated

In Walkthrough 1, we answered one country-level question:

> In Pakistan since 2000, which years should we flag because MCV1 coverage was
> low and measles incidence was elevated?

Today we ask the next manuscript question:

> Is Pakistan unusual, or do we see the same pattern across other countries in
> the region?

## What changes today?

The analysis rule does not change:

> Priority year = MCV1 coverage below 90% and at least 1 measles case per
> 100,000 people.

The workflow changes:

1. Write the rule once.
1. Test it on one country.
1. Repeat it across many countries.
1. Turn the repeated results into a table for the report.

## Load the data

```{r}
data_file <- file.path("data", "measles_final.Rds")

if (!file.exists(data_file)) {
  data_file <- file.path("..", "data", "measles_final.Rds")
}

measles <- readRDS(data_file)
```

## Before we loop

Loops are easier when we know the exact task.

For one country, we need to:

1. Keep rows for that country and year range.
1. Calculate cases per 100,000 people.
1. Apply the priority-year rule.
1. Count the priority years.
1. Return one row of results.

That is a function-shaped problem.

## Function 1: apply the rule

```{r}
add_priority_rule <- function(country_data,
                              coverage_cutoff = 90,
                              incidence_cutoff = 1) {
  country_data$cases_per_100k <- (
    country_data$measles_cases / country_data$population * 100000
  )

  country_data$priority_year <- ifelse(
    is.na(country_data$MCV1_coverage) | is.na(country_data$cases_per_100k),
    "missing information",
    ifelse(
      country_data$MCV1_coverage < coverage_cutoff &
        country_data$cases_per_100k >= incidence_cutoff,
      "priority year",
      "not priority"
    )
  )

  country_data
}
```

## Test the rule on one country

```{r}
pakistan <- measles[
  measles$country == "Pakistan" & measles$year >= 2000,
]

pakistan <- pakistan[order(pakistan$year), ]
pakistan <- add_priority_rule(pakistan)

tail(pakistan[, c("year", "MCV1_coverage", "cases_per_100k", "priority_year")])
```

## Function 2: summarize one country

```{r}
summarize_priority_years <- function(country_name,
                                     data,
                                     min_year = 2000,
                                     coverage_cutoff = 90,
                                     incidence_cutoff = 1) {
  country_data <- data[
    data$country == country_name & data$year >= min_year,
  ]

  country_data <- add_priority_rule(
    country_data, coverage_cutoff, incidence_cutoff
  )

  priority_years <- country_data$year[
    country_data$priority_year == "priority year"
  ]

  data.frame(
    country = country_name,
    region = country_data$region[1],
    sub_region = country_data$sub_region[1],
    years = nrow(country_data),
    priority_years = length(priority_years),
    missing_information_years = sum(country_data$priority_year == "missing information"),
    most_recent_priority_year = if (length(priority_years) == 0) {
      NA_integer_
    } else {
      max(priority_years)
    },
    max_cases_per_100k = if (all(is.na(country_data$cases_per_100k))) {
      NA_real_
    } else {
      round(max(country_data$cases_per_100k, na.rm = TRUE), 2)
    }
  )
}
```

## Test the summary function

```{r}
summarize_priority_years("Pakistan", data = measles)
```

## Why this helps

Before writing a loop, we made one repeatable unit:

```r
summarize_priority_years("Pakistan", data = measles)
```

If that works for Pakistan, the loop only needs to change the country name.

That is the main trick.

## Repeat with a `for` loop

```{r}
countries <- c("Afghanistan", "Bangladesh", "India", "Nepal", "Pakistan")

country_summaries <- data.frame()

for (country_name in countries) {
  one_summary <- summarize_priority_years(country_name, data = measles)
  country_summaries <- rbind(country_summaries, one_summary)
}
```

## Repeat with a `for` loop: result

```{r}
country_summaries[, c(
  "country",
  "priority_years",
  "most_recent_priority_year",
  "max_cases_per_100k"
)]
```

## What did the loop do?

Each loop pass did the same job:

1. Put one country name into `country_name`.
1. Run `summarize_priority_years()` for that country.
1. Add the one-row answer to the growing results table.

The computer is doing the repeated typing for us.

## Repeat with `lapply()`

The same idea can be written in a more functional style.

```{r}
summary_list <- lapply(
  countries,
  summarize_priority_years,
  data = measles
)

length(summary_list)
summary_list[[1]]
```

## Combine the list

```{r}
country_summaries_lapply <- do.call(rbind, summary_list)

country_summaries_lapply
```

## Which version should you use?

Use the version that helps you think clearly.

- A `for` loop is often easier to read while you are learning.
- `lapply()` is compact once the repeated task is already a function.
- Both are better than copying and pasting the same analysis five times.

## Now repeat over a whole region

```{r}
asia <- measles[measles$region == "Asia", ]
asian_countries <- sort(unique(asia$country))

head(asian_countries)
length(asian_countries)
```

## Run the country summary for Asia

```{r}
asia_summary_list <- lapply(
  asian_countries,
  summarize_priority_years,
  data = asia
)

asia_summaries <- do.call(rbind, asia_summary_list)
```

## Sort the results

```{r}
asia_summaries <- asia_summaries[
  order(-asia_summaries$priority_years, asia_summaries$country),
]
```

## Sort the results: top countries

```{r}
asia_top_columns <- asia_summaries[, c(
  "country",
  "priority_years",
  "most_recent_priority_year",
  "max_cases_per_100k"
)]

head(asia_top_columns, 5)
```

## Sort the results: top countries, continued

```{r}
head(asia_top_columns, 10)[6:10, ]
```

## Make a manuscript-style table

```{r}
top_asia <- head(asia_summaries, 10)

top_asia_table <- top_asia[, c(
  "country",
  "sub_region",
  "priority_years",
  "missing_information_years",
  "most_recent_priority_year",
  "max_cases_per_100k"
)]

top_asia_counts <- top_asia_table[, c(
  "country",
  "sub_region",
  "priority_years"
)]

top_asia_details <- top_asia_table[, c(
  "country",
  "missing_information_years",
  "most_recent_priority_year",
  "max_cases_per_100k"
)]
```

## Make a manuscript-style table: counts

```{r}
head(top_asia_counts, 5)
```

## Make a manuscript-style table: counts, continued

```{r}
tail(top_asia_counts, 5)
```

## Make a manuscript-style table: details

```{r}
head(top_asia_details, 5)
```

## Make a manuscript-style table: details, continued

```{r}
tail(top_asia_details, 5)
```

## Make a quick figure

```{r}
barplot(
  top_asia$priority_years,
  names.arg = top_asia$country,
  las = 2,
  ylab = "Priority years since 2000",
  main = "Countries in Asia with the most priority years"
)
```

## What goes in the report?

**Methods sentence**

We wrote a reusable function to classify each country-year as a priority year
when MCV1 coverage was below 90% and measles incidence was at least 1 case per
100,000 people, then applied that function to every country in Asia.

**Results sentence**

The repeated analysis identified several countries in Asia with many priority
years since 2000, suggesting that the Pakistan pattern was not unique under
this simple rule.

## Active checkpoint

**Student prompt**

Repeat the regional analysis for Africa.

Your goal is to produce a table of the 10 countries in Africa with the most
priority years since 2000.

## Active checkpoint: hint

Reuse the objects we already built.

You only need to change:

1. the region filter
1. the vector of country names
1. the object names, if you want them to be easier to read

## Active checkpoint: solution

```{r}
africa <- measles[measles$region == "Africa", ]
african_countries <- sort(unique(africa$country))

africa_summary_list <- lapply(
  african_countries,
  summarize_priority_years,
  data = africa
)

africa_summaries <- do.call(rbind, africa_summary_list)
```

## Active checkpoint: solution sort

```{r}
africa_summaries <- africa_summaries[
  order(-africa_summaries$priority_years, africa_summaries$country),
]
```

## Active checkpoint: solution table

```{r}
africa_top_columns <- africa_summaries[, c(
  "country",
  "priority_years"
)]

head(africa_top_columns, 5)
```

## Active checkpoint: solution table, continued

```{r}
head(africa_top_columns, 10)[6:10, ]
```

## Active checkpoint: one more change

What if the manuscript team decides that 1 case per 100,000 is too sensitive?

Try the Africa analysis again with:

- MCV1 cutoff: 90%
- incidence cutoff: 5 cases per 100,000

## Active checkpoint: solution

```{r}
africa_summary_list_strict <- lapply(
  african_countries,
  summarize_priority_years,
  data = africa,
  coverage_cutoff = 90,
  incidence_cutoff = 5
)

africa_summaries_strict <- do.call(rbind, africa_summary_list_strict)
```

## Active checkpoint: strict solution sort

```{r}
africa_summaries_strict <- africa_summaries_strict[
  order(-africa_summaries_strict$priority_years, africa_summaries_strict$country),
]
```

## Active checkpoint: strict solution table

```{r}
africa_strict_top_columns <- africa_summaries_strict[, c(
  "country",
  "priority_years"
)]

head(africa_strict_top_columns, 5)
```

## Active checkpoint: strict solution table, continued

```{r}
head(africa_strict_top_columns, 10)[6:10, ]
```

## What did we practice?

- Turning one successful analysis into a function
- Repeating that function with a `for` loop
- Repeating that function with `lapply()`
- Combining repeated results with `do.call(rbind, ...)`
- Creating a manuscript-ready summary table from many repeated analyses

In Exercise 2, this same habit helps you move from a draft analysis to a
polished report.
