---
title: "Data science walkthrough 1"
format:
  revealjs:
    toc: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## An example research question

> Since 2000, did the country in Asia with the lowest MCV1 coverage also have the highest measles incidence per capita?

## The programming idea

- This is something that would be pretty easy to do by hand (e.g. with Excel), since the dataset is not all that big.
- The difficult part for programming is **translating this question into computer code.**
- The computer will always do exactly what we tell it to, so we have to turn the steps to solving this question into specific computer code.

## Steps to solve this problem

1. Get the dataset
1. Separate out countries in Asia and years since 2000
1. Calculate per capita incidence
1. Find the country with the lowest MCV1 coverage
1. Find the country with the highest per capita incidence
1. Are they the same?

## Load the data

```{r}
data_file <- here::here("data", "measles_final.Rds")

measles <- readr::read_rds(data_file)
```

## Look at the data

```{r}
str(measles)
```

## What do we need?

For this question, we need:

- `country`, because we need to compare the countries in the two results
- `region`, because we only want countries in Asia
- `year`, because we only want 2000 and later
- `measles_cases`, because we need incidence
- `population`, because raw case counts are hard to compare
- `MCV1_coverage`, because that is our coverage variable

## Step 1: keep the country-years we need

```{r}
asia2000 <- subset(measles, region == "Asia" & year >= 2000)
```

## Check the subset

```{r}
head(asia2000)
```

## Check the subset

```{r}
summary(asia2000)
```

## Step 2: calculate incidence

::: {.incremental}
* Raw case counts are useful, but they do not account for population size. Any ideas?
* cases per 100,000 people
:::

. . .

```{r}
asia2000$cases_per_100k <- asia2000$measles_cases / asia2000$population * 100000
```

## Check the calculation

```{r}
summary(asia2000)
```

## Step 3: lowest coverage

Now we need to find the row with the lowest MCV1 coverage.

`which.min()` gives us the row number for the smallest value in a vector.

```{r}
lowest_coverage_row <- asia2000[which.min(asia2000$MCV1_coverage), c(
  "country",
  "year",
  "MCV1_coverage",
  "measles_cases",
  "population",
  "cases_per_100k"
)]

lowest_coverage_row
```

## What did `which.min()` do?

::: {.incremental}
* `asia2000$MCV1_coverage` is one long vector of coverage values.
* `which.min(asia2000$MCV1_coverage)` finds the position of the smallest non-missing value.
* We use that position to pull out the whole row.
:::

## Step 4: highest incidence

Now we do the same thing for measles incidence.

This time we want the largest value, so we use `which.max()`.

```{r}
highest_incidence_row <- asia2000[which.max(asia2000$cases_per_100k), c(
  "country",
  "year",
  "MCV1_coverage",
  "measles_cases",
  "population",
  "cases_per_100k"
)]

highest_incidence_row
```

## Step 5: are they the same country?

We now have two rows:

1. the row with the lowest MCV1 coverage
1. the row with the highest measles incidence per capita

The research question asks whether the **country** is the same.

```{r}
lowest_coverage_country <- lowest_coverage_row$country
highest_incidence_country <- highest_incidence_row$country

same_country <- lowest_coverage_country == highest_incidence_country

lowest_coverage_country
highest_incidence_country
```

## Step 6: write the answer with `if` {.scrollable}

This is a different kind of conditional.

Here we are not labeling every row. We are asking one yes/no question and then
choosing one sentence.

```{r}
if (same_country) {
  paste0(
    "Yes. ",
    lowest_coverage_country,
    " had both the lowest MCV1 coverage and the highest measles incidence ",
    "per capita."
  )
} else {
  paste0(
    "No. ",
    lowest_coverage_country,
    " had the lowest MCV1 coverage, but ",
    highest_incidence_country,
    " had the highest measles incidence per capita."
  )
}
```

## Make a small table

```{r}
answer_table <- data.frame(
  statistic = c("Lowest MCV1 coverage", "Highest measles incidence"),
  country = c(lowest_coverage_row$country, highest_incidence_row$country),
  year = c(lowest_coverage_row$year, highest_incidence_row$year),
  MCV1_coverage = c(
    lowest_coverage_row$MCV1_coverage,
    highest_incidence_row$MCV1_coverage
  ),
  cases_per_100k = c(
    lowest_coverage_row$cases_per_100k,
    highest_incidence_row$cases_per_100k
  )
)

answer_table
```

## Make a quick figure {.scrollable}

```{r}
plot_rows <- !is.na(asia2000$MCV1_coverage) &
  !is.na(asia2000$cases_per_100k)

plot(
  asia2000$MCV1_coverage[plot_rows],
  asia2000$cases_per_100k[plot_rows],
  pch = 16,
  col = "gray70",
  xlab = "MCV1 coverage (%)",
  ylab = "Measles cases per 100,000 people",
  main = "Asian country-years since 2000"
)

points(
  lowest_coverage_row$MCV1_coverage,
  lowest_coverage_row$cases_per_100k,
  pch = 19,
  col = "red",
  cex = 1.5
)

points(
  highest_incidence_row$MCV1_coverage,
  highest_incidence_row$cases_per_100k,
  pch = 19,
  col = "blue",
  cex = 1.5
)

legend(
  "topright",
  legend = c("Lowest MCV1 coverage", "Highest incidence"),
  pch = 19,
  col = c("red", "blue"),
  bty = "n"
)
```

## You try it!

Change the question:

> Since 2010, did the country in the Americas with the lowest MCV2 coverage also
> have the highest measles incidence per capita?

Write the code before looking at the solution.

## You try it!: hint

You need to change four pieces:

1. the region
1. the start year
1. the coverage variable
1. the object names, so you do not overwrite the first answer

## You try it!: solution

```{r}
americas2010 <- subset(measles, region == "Americas" & year >= 2010)

americas2010$cases_per_100k <- americas2010$measles_cases /
  americas2010$population * 100000
```

## You try it!: solution rows

```{r}
lowest_mcv2_row <- americas2010[which.min(americas2010$MCV2_coverage), c(
  "country",
  "year",
  "MCV2_coverage",
  "measles_cases",
  "population",
  "cases_per_100k"
)]

highest_americas_incidence_row <- americas2010[
  which.max(americas2010$cases_per_100k),
  c(
    "country",
    "year",
    "MCV2_coverage",
    "measles_cases",
    "population",
    "cases_per_100k"
  )
]

lowest_mcv2_row$cases_per_100k <- round(
  lowest_mcv2_row$cases_per_100k,
  2
)

highest_americas_incidence_row$cases_per_100k <- round(
  highest_americas_incidence_row$cases_per_100k,
  2
)

lowest_mcv2_row
```

## You try it!: solution rows, continued

```{r}
highest_americas_incidence_row
```

## You try it!: compare them

```{r}
lowest_mcv2_row$country
highest_americas_incidence_row$country
```

## What did we practice?

- Starting with a question that a reader can understand
- Filtering rows with a logical condition
- Creating an incidence rate
- Finding minimum and maximum rows with `which.min()` and `which.max()`
- Using `if` and `else` to turn a comparison into a sentence
- Creating a table, figure, and short manuscript-style interpretation

Next time, we will repeat ideas like this across many countries.
