---
title: "Data science walkthrough 1: Conditionals"
format:
  revealjs:
    toc: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## A small manuscript problem

We are going to answer one concrete question with the measles data:

> In Pakistan since 2000, which years should we flag for follow-up because first-dose measles vaccine coverage was low and measles incidence was elevated?

## A small manuscript problem: manuscript arc

This is a miniature version of the Exercise 1 manuscript arc:

1. State a question.
1. Translate the question into a data rule.
1. Run the rule.
1. Write a cautious result.

## The programming idea

We can describe the rule in plain English:

> If MCV1 coverage is below 90% and measles incidence is at least 1 case per 100,000 people, call it a priority year.

Now the problem is not "Can I memorize R?"

The problem is:

> How do I turn that sentence into computer code?

## Load the data

```{r}
data_file <- file.path("data", "measles_final.Rds")

if (!file.exists(data_file)) {
  data_file <- file.path("..", "data", "measles_final.Rds")
}

measles <- readRDS(data_file)
```

## Look at the data

```{r}
str(measles[, c(
  "country",
  "year",
  "measles_cases",
  "population",
  "MCV1_coverage"
)])
```

## What do we need?

For this question, we need:

- `country`, because we are starting with Pakistan
- `year`, because we only want 2000 and later
- `measles_cases`, because we need incidence
- `population`, because raw case counts are hard to compare
- `MCV1_coverage`, because that is our coverage variable

## Step 1: keep the country-years we need

```{r}
pakistan <- measles[
  measles$country == "Pakistan" & measles$year >= 2000,
]

pakistan <- pakistan[order(pakistan$year), ]
```

## Check the subset

```{r}
pakistan_check_rows <- pakistan[, c(
  "country",
  "year",
  "measles_cases",
  "MCV1_coverage"
)]

head(pakistan_check_rows)
```

## Check the subset, continued

```{r}
tail(pakistan_check_rows)
```

## Step 2: calculate incidence

Raw case counts are useful, but they do not account for population size.

So we make a new variable:

> measles cases per 100,000 people

```{r}
pakistan$cases_per_100k <- pakistan$measles_cases / pakistan$population * 100000
```

## Check the calculation

```{r}
pakistan_check <- pakistan[, c(
  "year",
  "measles_cases",
  "population",
  "cases_per_100k"
)]

pakistan_check$cases_per_100k <- round(pakistan_check$cases_per_100k, 2)

tail(pakistan_check, 8)
```

## Step 3: translate each condition

The English rule has two parts:

1. MCV1 coverage is below 90%.
1. Measles incidence is at least 1 case per 100,000 people.

In R, each part becomes a logical vector.

```{r}
low_coverage <- pakistan$MCV1_coverage < 90
elevated_incidence <- pakistan$cases_per_100k >= 1
```

## Inspect the logical vectors

```{r}
logic_check <- data.frame(
  year = pakistan$year,
  MCV1_coverage = pakistan$MCV1_coverage,
  cases_per_100k = round(pakistan$cases_per_100k, 2),
  low_coverage = low_coverage,
  elevated_incidence = elevated_incidence
)

tail(logic_check, 8)
```

## Missing values matter

If coverage is missing, R cannot decide whether coverage is below 90%.

```{r}
table(low_coverage, useNA = "ifany")
```

That is not a failure. It is R refusing to guess.

For a manuscript-style analysis, we need to decide what we want to do with
missing information.

## Step 4: write the decision rule

Here is the rule in full:

1. If coverage or incidence is missing, label the row `"missing information"`.
1. Otherwise, if coverage is below 90% and incidence is at least 1, label the
   row `"priority year"`.
1. Otherwise, label the row `"not priority"`.

## Step 4: code the decision rule

```{r}
pakistan$priority_year <- ifelse(
  is.na(pakistan$MCV1_coverage) | is.na(pakistan$cases_per_100k),
  "missing information",
  ifelse(
    pakistan$MCV1_coverage < 90 & pakistan$cases_per_100k >= 1,
    "priority year",
    "not priority"
  )
)
```

## What did `ifelse()` do?

`ifelse()` has three pieces:

```r
ifelse(test, value_if_true, value_if_false)
```

Our nested version says:

- first check whether the data are missing
- if not, check whether the year meets the priority rule
- give every row one label

## Step 5: inspect the answer

```{r}
pakistan_result <- pakistan[, c(
  "year",
  "measles_cases",
  "MCV1_coverage",
  "cases_per_100k",
  "priority_year"
)]

pakistan_result$cases_per_100k <- round(pakistan_result$cases_per_100k, 2)
```

## Step 5: inspect recent years

```{r}
tail(pakistan_result, 8)
```

## Step 6: summarize the answer

```{r}
table(pakistan$priority_year)
```

```{r}
pakistan$year[pakistan$priority_year == "priority year"]
```

## Make a small manuscript table

```{r}
priority_years <- pakistan$year[pakistan$priority_year == "priority year"]

data.frame(
  country = "Pakistan",
  start_year = min(pakistan$year),
  end_year = max(pakistan$year),
  total_years = nrow(pakistan),
  priority_years = length(priority_years),
  most_recent_priority_year = max(priority_years)
)
```

## Make a quick figure

```{r}
plot(
  pakistan$year,
  pakistan$cases_per_100k,
  type = "b",
  xlab = "Year",
  ylab = "Measles cases per 100,000 people",
  main = "Pakistan measles incidence since 2000"
)

abline(h = 1, lty = 2)

points(
  pakistan$year[pakistan$priority_year == "priority year"],
  pakistan$cases_per_100k[pakistan$priority_year == "priority year"],
  pch = 19,
  col = "red"
)
```

## What goes in the report?

**Methods sentence**

We defined a priority year as a country-year with MCV1 coverage below 90% and
at least 1 reported measles case per 100,000 people; country-years with missing
coverage or incidence were labeled as missing information.

**Results sentence**

Pakistan had several priority years since 2000 under this rule, with the most
recent priority year identified in the summary table.

## Active checkpoint

**Student prompt**

Change the question:

> In Afghanistan since 2010, which years should be flagged if MCV2 coverage was
> below 80% or measles incidence was at least 5 cases per 100,000 people?

Write the code before looking at the solution.

## Active checkpoint: hint

You need to change four pieces:

1. the country
1. the start year
1. the coverage variable and cutoff
1. the logical connector, because this rule uses "or"

The R operator for "or" is `|`.

## Active checkpoint: solution

```{r}
afghanistan <- measles[
  measles$country == "Afghanistan" & measles$year >= 2010,
]

afghanistan <- afghanistan[order(afghanistan$year), ]
afghanistan$cases_per_100k <- afghanistan$measles_cases / afghanistan$population * 100000
```

## Active checkpoint: solution rule

```{r}
afghanistan$priority_year <- ifelse(
  (!is.na(afghanistan$MCV2_coverage) & afghanistan$MCV2_coverage < 80) |
    (!is.na(afghanistan$cases_per_100k) & afghanistan$cases_per_100k >= 5),
  "priority year",
  ifelse(
    is.na(afghanistan$MCV2_coverage) | is.na(afghanistan$cases_per_100k),
    "missing information",
    "not priority"
  )
)
```

## Active checkpoint: solution table

```{r}
afghanistan_result <- afghanistan[, c(
  "year",
  "measles_cases",
  "MCV2_coverage",
  "cases_per_100k",
  "priority_year"
)]

afghanistan_result$cases_per_100k <- round(afghanistan_result$cases_per_100k, 2)

afghanistan_labels <- afghanistan_result[, c("year", "priority_year")]
head(afghanistan_labels, 7)
```

## Active checkpoint: solution table, continued

```{r}
tail(afghanistan_labels, 7)
```

## Active checkpoint: summarize it

```{r}
table(afghanistan$priority_year)
```

```{r}
afghanistan$year[afghanistan$priority_year == "priority year"]
```

## What did we practice?

- Starting with a question that a reader can understand
- Turning a plain-English rule into logical conditions
- Using `&`, `|`, `is.na()`, and `ifelse()`
- Creating a table, figure, and short manuscript-style interpretation

Next time, we will repeat this same rule across many countries.
