---
title: "Data science walkthrough 2"
format:
  revealjs:
    toc: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## From one region to many countries

Last time we looked at coverage and incidence in one specific region.

That works when the question is small enough to inspect directly.

This time, let's think about a bigger manuscript problem:

> How can we flag issues across multiple countries when there is too much data
> to look through manually?

## What changes today?

To make the repeated task concrete, we will use a simple flagging rule:

> Priority year = MCV1 coverage below 90% and at least 1 measles case per
> 100,000 people.

The workflow changes:

1. Write the rule once.
1. Test it on one country.
1. Repeat it across many countries.
1. Turn the repeated results into a table for the report.

## Load the data

```{r}
data_file <- here::here("data", "measles_final.Rds")

measles <- readr::read_rds(data_file)
```

## Before we loop

Loops are easier when we know the exact task.

For one country, we need to:

1. Keep rows for that country and year range.
1. Calculate cases per 100,000 people.
1. Apply the priority-year rule.
1. Count the priority years.
1. Return one row of results.

. . .

Sounds like a good use of a function!

## Function 1: apply the rule

```{r}
add_priority_rule <- function(country_data,
                              coverage_cutoff = 90,
                              incidence_cutoff = 1) {
  country_data$cases_per_100k <- (
    country_data$measles_cases / country_data$population * 100000
  )
  
  mcv_coverage_flag <- country_data$MCV1_coverage < coverage_cutoff
  
  incidence_flag <- country_data$cases_per_100k >= incidence_cutoff

  country_data$priority_year <- mcv_coverage_flag & incidence_flag

  return(country_data)
}
```

## Test the rule on one country

```{r}
pakistan <- measles[
  measles$country == "Pakistan" & measles$year >= 2000,
]

pakistan <- pakistan[order(pakistan$year), ]
pakistan <- add_priority_rule(pakistan)

head(pakistan[, c("year", "MCV1_coverage", "cases_per_100k", "priority_year")])
```

## Function 2: summarize one country

```{r}
summarize_priority_years <- function(country_name,
                                     data,
                                     min_year = 2000,
                                     coverage_cutoff = 90,
                                     incidence_cutoff = 1) {
  country_data <- subset(
    data,
    country == country_name & year >= min_year
  )

  country_data <- add_priority_rule(
    country_data, coverage_cutoff, incidence_cutoff
  )

  priority_years <- subset(country_data, priority_year)$year

  data.frame(
    country = country_name,
    region = country_data$region[1],
    sub_region = country_data$sub_region[1],
    years = nrow(country_data),
    priority_years = length(priority_years),
    most_recent_priority_year = if (length(priority_years) == 0) {
      NA_integer_
    } else {
      max(priority_years)
    },
    max_cases_per_100k = if (all(is.na(country_data$cases_per_100k))) {
      NA_real_
    } else {
      round(max(country_data$cases_per_100k, na.rm = TRUE), 2)
    }
  )
}
```

## Test the summary function

```{r}
summarize_priority_years("Pakistan", data = measles)
```

## Why this helps

Before writing a loop, we made one repeatable unit:

```r
summarize_priority_years("Pakistan", data = measles)
```

If that works for Pakistan, the loop only needs to change the country name.

## Repeat with a `for` loop

```{r}
countries <- c("Afghanistan", "Bangladesh", "India", "Nepal", "Pakistan")

country_summaries <- data.frame()

for (country_name in countries) {
  one_summary <- summarize_priority_years(country_name, data = measles)
  country_summaries <- rbind(country_summaries, one_summary)
}
```

## Repeat with a `for` loop: result

```{r}
country_summaries
```

## What did the loop do?

Each loop pass did the same job:

1. Put one country name into `country_name`.
1. Run `summarize_priority_years()` for that country.
1. Add the one-row answer to the growing results table.

## Repeat with `lapply()`

The same idea can be written in a more functional style.

```{r}
summary_list <- lapply(
  countries,
  summarize_priority_years,
  data = measles
)

length(summary_list)
summary_list[[1]]
```

## Combine the list

- It's worth understanding how to use `do.call()` even though it's weird.

```{r}
country_summaries_lapply <- do.call(rbind, summary_list)

country_summaries_lapply
```

## Which version should you use?

Use the version that helps you think clearly.

- A `for` loop is often easier to read while you are learning.
- `lapply()` is compact once the repeated task is already a function.
- Both are better than copying and pasting the same analysis five times.

## Now repeat over a whole region

```{r}
asia <- measles[measles$region == "Asia", ]
asian_countries <- sort(unique(asia$country))

head(asian_countries)
length(asian_countries)
```

## Run the country summary for Asia

```{r}
asia_summary_list <- lapply(
  asian_countries,
  summarize_priority_years,
  data = asia
)

asia_summaries <- do.call(rbind, asia_summary_list)
```

## Sort the results

```{r}
asia_summaries <- asia_summaries[
  order(-asia_summaries$priority_years, asia_summaries$country),
]

head(asia_summaries)
```

## Make a manuscript-style table

```{r}
top_asia <- head(asia_summaries, 10)

top_asia_table <- top_asia[, c(
  "country",
  "sub_region",
  "priority_years",
  "most_recent_priority_year",
  "max_cases_per_100k"
)]

top_asia_counts <- top_asia_table[, c(
  "country",
  "sub_region",
  "priority_years"
)]

top_asia_details <- top_asia_table[, c(
  "country",
  "most_recent_priority_year",
  "max_cases_per_100k"
)]
```

## Make a manuscript-style table: counts

```{r}
head(top_asia_counts, 5)
```

## Make a manuscript-style table: counts, continued

```{r}
tail(top_asia_counts, 5)
```

## Make a manuscript-style table: details

```{r}
head(top_asia_details, 5)
```

## Make a quick figure

```{r}
barplot(
  top_asia$priority_years,
  names.arg = top_asia$country,
  las = 2,
  ylab = "Priority years since 2000",
  main = "Countries in Asia with the most priority years"
)
```

## You try it!

**Student prompt**

Repeat the regional analysis for Africa.

Your goal is to produce a table of the 10 countries in Africa with the most
priority years since 2000.

## You try it!: hint

Reuse the objects we already built.

You only need to change:

1. the region filter
1. the vector of country names
1. the object names, if you want them to be easier to read

## You try it!: solution

```{r}
africa <- measles[measles$region == "Africa", ]
african_countries <- sort(unique(africa$country))

africa_summary_list <- lapply(
  african_countries,
  summarize_priority_years,
  data = africa
)

africa_summaries <- do.call(rbind, africa_summary_list)
```

## You try it!: solution sort

```{r}
africa_summaries <- africa_summaries[
  order(-africa_summaries$priority_years, africa_summaries$country),
]
```

## You try it!: solution table

```{r}
africa_top_columns <- africa_summaries[, c(
  "country",
  "priority_years"
)]

head(africa_top_columns, 5)
```

## You try it!: solution table, continued

```{r}
head(africa_top_columns, 10)[6:10, ]
```

## You try it!: one more change

What if the manuscript team decides that 1 case per 100,000 is too sensitive?

Try the Africa analysis again with:

- MCV1 cutoff: 90%
- incidence cutoff: 5 cases per 100,000

## You try it!: solution

```{r}
africa_summary_list_strict <- lapply(
  african_countries,
  summarize_priority_years,
  data = africa,
  coverage_cutoff = 90,
  incidence_cutoff = 5
)

africa_summaries_strict <- do.call(rbind, africa_summary_list_strict)
```

## You try it!: strict solution sort

```{r}
africa_summaries_strict <- africa_summaries_strict[
  order(-africa_summaries_strict$priority_years, africa_summaries_strict$country),
]
```

## You try it!: strict solution table

```{r}
africa_strict_top_columns <- africa_summaries_strict[, c(
  "country",
  "priority_years"
)]

head(africa_strict_top_columns, 5)
```

## You try it!: strict solution table, continued

```{r}
head(africa_strict_top_columns, 10)[6:10, ]
```

## What did we practice?

- Turning one successful analysis into a function
- Repeating that function with a `for` loop
- Repeating that function with `lapply()`
- Combining repeated results with `do.call(rbind, ...)`
- Creating a manuscript-ready summary table from many repeated analyses

In Exercise 2, this same habit helps you move from a draft analysis to a
polished report.
