---
title: "Exercise 1 Solutions: Start a QCRC Mini-Manuscript"
format: html
editor: visual
execute:
  warning: false
  message: false
---

# Instructor note

This is one acceptable solution path for Exercise 1. It uses the recommended
QCRC 30-day mortality question. Student reports may choose different exposures
or a different course dataset if they preserve the required deliverables.

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(here)
library(janitor)
library(knitr)
library(readxl)
```

# Task 1 solution: Create a report draft

Students should start from `exercises/qcrc-mini-manuscript-starter.qmd`, save a
working copy, and render it. The starter file is already a valid solution
skeleton because it contains Introduction, Methods, Results, Discussion, and
Appendix sections.

# Task 2 solution: Load and inspect QCRC

```{r}
#| label: load-clean-data

to_numeric <- function(x) {
  as.numeric(dplyr::na_if(as.character(x), "."))
}

qcrc <- readxl::read_xlsx(
  here::here("data", "QCRC_FINAL_Deidentified.xlsx"),
  sheet = "Main_Dataset"
) |>
  janitor::clean_names() |>
  dplyr::mutate(
    bmi = to_numeric(bmi),
    crp = to_numeric(crp),
    d_dimer = to_numeric(d_dimer),
    x30d_mortality = factor(
      x30d_mortality,
      levels = c(0, 1),
      labels = c("No 30-day mortality", "30-day mortality")
    ),
    intubated = factor(intubated, levels = c(0, 1), labels = c("No", "Yes")),
    remdesivir_or_placebo = factor(
      remdesivir_or_placebo,
      levels = c(0, 1),
      labels = c("Placebo", "Remdesivir")
    ),
    female = factor(female),
    race = factor(race)
  )

data.frame(rows = nrow(qcrc), columns = ncol(qcrc)) |>
  knitr::kable(caption = "QCRC main dataset dimensions.")

names(qcrc)[1:10]
```

```{r}
#| label: missingness

analysis_vars <- c(
  "x30d_mortality",
  "age",
  "bmi",
  "female",
  "intubated",
  "remdesivir_or_placebo"
)

missing_counts <- sapply(qcrc[analysis_vars], \(x) sum(is.na(x)))

data.frame(
  variable = names(missing_counts),
  missing = as.integer(missing_counts)
) |>
  knitr::kable(caption = "Missingness check for Exercise 1 variables.")
```

Example Methods sentence:

> We analyzed the `Main_Dataset` sheet from the deidentified QCRC COVID-19 ICU
> workbook. We cleaned variable names, converted selected lab and BMI variables
> from text to numeric values, and recoded binary variables as factors for
> reporting.

# Task 3 solution: Define the manuscript variables

Example introduction:

> Severe COVID-19 outcomes vary across patient characteristics and clinical
> factors. In this report, we ask whether age, BMI, intubation, and treatment
> group are associated with 30-day mortality among patients in the QCRC ICU
> dataset. We hypothesize that older age and intubation will be associated with
> higher odds of 30-day mortality. This analysis is descriptive and
> associational; it should not be interpreted as proving causality.

# Task 4 solution: Descriptive table or figure

```{r}
#| label: descriptive-table

qcrc |>
  dplyr::group_by(x30d_mortality) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_age = round(mean(age, na.rm = TRUE), 1),
    mean_bmi = round(mean(bmi, na.rm = TRUE), 1),
    intubated_percent = round(mean(intubated == "Yes", na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Selected characteristics by 30-day mortality status.")
```

```{r}
#| label: bmi-figure
#| fig-cap: "BMI distribution by 30-day mortality status in the QCRC dataset."

ggplot(qcrc, aes(x = x30d_mortality, y = bmi)) +
  geom_boxplot(na.rm = TRUE) +
  labs(x = NULL, y = "BMI")
```

Example interpretation:

> Patients with and without 30-day mortality can be compared on age, BMI, and
> intubation status before fitting a model. The descriptive table is not an
> adjusted analysis, but it helps identify patterns and missingness that should
> be considered in Exercise 2.

# Task 5 solution: First-pass analysis

```{r}
#| label: first-pass-test

t.test(bmi ~ x30d_mortality, data = qcrc)
```

```{r}
#| label: first-pass-model

model_30d <- glm(
  x30d_mortality ~ age + bmi + intubated + remdesivir_or_placebo,
  data = qcrc,
  family = binomial
)

summary(model_30d)
```

```{r}
#| label: odds-ratio-table

or_table <- data.frame(
  term = names(coef(model_30d)),
  odds_ratio = exp(coef(model_30d)),
  exp(confint.default(model_30d)),
  check.names = FALSE
)

names(or_table) <- c("term", "odds_ratio", "conf_low", "conf_high")

or_table |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "First-pass adjusted odds ratios.")
```

Example interpretation:

> The logistic regression estimates adjusted associations with 30-day
> mortality. Because this is an observational-style course exercise with limited
> cleaning and variable review, the estimates should be treated as a first-pass
> signal rather than a definitive clinical conclusion.

# Task 6 solution: Mini-manuscript draft

Expected draft structure:

- Introduction: one paragraph with question and hypothesis.
- Methods: data source, variable cleaning, and analysis choices.
- Results: one descriptive table or figure, one test or model, and a short
  interpretation.
- Discussion: main takeaway, limitation, and next step for Exercise 2.

# Advanced task solution outlines

1. Workbook inventory: use `readxl::excel_sheets()`, `lapply()`, `nrow()`, and
   `ncol()` to build a sheet-level table.
1. Helper summary function: write `summarize_numeric <- function(data, var)` or
   use a vector input such as `summarize_numeric <- function(x)`.
1. Sensitivity endpoint: refit the same code with `died` or `x60d_mortality`
   after recoding that endpoint as a factor.
1. Alternate dataset: require the same question, data, descriptive result,
   analysis, interpretation, and limitation structure.
