---
title: "Exercise 2 Solutions: Polish the QCRC Report"
format: html
editor: visual
execute:
  warning: false
  message: false
---

# Instructor note

This solution continues the Exercise 1 QCRC report. It is an exemplar structure,
not the only valid student answer.

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(here)
library(janitor)
library(knitr)
library(readxl)
```

# Task 1 solution: Re-render and triage

Example Day 2 triage list:

- Reproducibility fix: move data loading and recoding into a named function.
- Analysis fix: report both crude and adjusted logistic models for 30-day
  mortality.
- Formatting fix: replace raw model output with a compact odds-ratio table and
  add captions.

# Task 2 solution: Reusable data loading

```{r}
#| label: reusable-loader

to_numeric <- function(x) {
  as.numeric(dplyr::na_if(as.character(x), "."))
}

read_qcrc_main <- function(path = here::here("data", "QCRC_FINAL_Deidentified.xlsx")) {
  readxl::read_xlsx(path, sheet = "Main_Dataset") |>
    janitor::clean_names() |>
    dplyr::mutate(
      bmi = to_numeric(bmi),
      crp = to_numeric(crp),
      d_dimer = to_numeric(d_dimer),
      vent_los = to_numeric(vent_los),
      x30d_mortality = factor(
        x30d_mortality,
        levels = c(0, 1),
        labels = c("No 30-day mortality", "30-day mortality")
      ),
      died = factor(died, levels = c(0, 1), labels = c("Survived", "Died")),
      intubated = factor(intubated, levels = c(0, 1), labels = c("No", "Yes")),
      remdesivir_or_placebo = factor(
        remdesivir_or_placebo,
        levels = c(0, 1),
        labels = c("Placebo", "Remdesivir")
      ),
      female = factor(female),
      race = factor(race)
    )
}

qcrc <- read_qcrc_main()
```

Example Methods sentence:

> We used a report-local helper function to read the QCRC main sheet, clean
> variable names, convert selected character variables to numeric values, and
> recode binary variables for analysis and reporting.

# Task 3 solution: Workbook inventory

```{r}
#| label: workbook-inventory

data_file <- here::here("data", "QCRC_FINAL_Deidentified.xlsx")
sheet_names <- readxl::excel_sheets(data_file)
sheet_list <- lapply(sheet_names, \(s) readxl::read_xlsx(data_file, sheet = s))

workbook_inventory <- data.frame(
  sheet = sheet_names,
  rows = sapply(sheet_list, nrow),
  columns = sapply(sheet_list, ncol)
)

workbook_inventory |>
  knitr::kable(caption = "QCRC workbook inventory.")
```

# Task 4 solution: Analysis table and figure

```{r}
#| label: final-descriptive-table

qcrc |>
  dplyr::group_by(x30d_mortality) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_age = round(mean(age, na.rm = TRUE), 1),
    median_icu_los = round(median(icu_los, na.rm = TRUE), 1),
    mean_bmi = round(mean(bmi, na.rm = TRUE), 1),
    intubated_percent = round(mean(intubated == "Yes", na.rm = TRUE) * 100, 1),
    remdesivir_percent = round(
      mean(remdesivir_or_placebo == "Remdesivir", na.rm = TRUE) * 100,
      1
    ),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Selected characteristics by 30-day mortality status.")
```

```{r}
#| label: age-mortality-figure
#| fig-cap: "Age distribution by 30-day mortality status."

ggplot(qcrc, aes(x = x30d_mortality, y = age)) +
  geom_boxplot(na.rm = TRUE) +
  labs(x = NULL, y = "Age")
```

Example interpretation:

> The descriptive table and age figure help establish the sample context before
> adjusted modeling. Differences in descriptive summaries should not be treated
> as adjusted effects.

# Task 5 solution: Crude and adjusted models

```{r}
#| label: crude-adjusted-models

crude_model <- glm(
  x30d_mortality ~ intubated,
  data = qcrc,
  family = binomial
)

adjusted_model <- glm(
  x30d_mortality ~ intubated + age + bmi + remdesivir_or_placebo,
  data = qcrc,
  family = binomial
)

summary(adjusted_model)
```

```{r}
#| label: model-table-function

tidy_or <- function(model) {
  out <- data.frame(
    term = names(coef(model)),
    odds_ratio = exp(coef(model)),
    exp(confint.default(model)),
    check.names = FALSE
  )

  names(out) <- c("term", "odds_ratio", "conf_low", "conf_high")

  out |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2)))
}

dplyr::bind_rows(
  dplyr::mutate(tidy_or(crude_model), model = "Crude"),
  dplyr::mutate(tidy_or(adjusted_model), model = "Adjusted")
) |>
  dplyr::select(model, term, odds_ratio, conf_low, conf_high) |>
  knitr::kable(caption = "Crude and adjusted odds ratios for 30-day mortality.")
```

Example Results paragraph:

> In the crude model, intubation was associated with higher odds of 30-day
> mortality. After adjustment for age, BMI, and treatment group, the association
> remained positive, although the exact estimate and confidence interval should
> be interpreted in light of missingness, variable coding, and the limited
> scope of this course exercise.

# Task 6 solution: Polished deliverable

Expected final report structure:

- Title: specific and outcome-oriented.
- Introduction: question, rationale, and hypothesis.
- Methods: data source, cleaning, descriptive analysis, model formula, and
  uncertainty approach.
- Results: table, figure, model table, and short interpretation.
- Discussion: main finding, practical meaning, limitation, and next step.
- Appendix: workbook inventory and exploratory checks.

# Advanced task solution outlines

1. Model comparison: fit two nested models, then compare with `AIC(model1,
   model2)` or `anova(model1, model2, test = "Chisq")`; remind students this
   does not choose confounders causally.
1. Predictions: create a small `newdata` grid and call `predict(model, newdata,
   type = "response", se.fit = TRUE)`; report predicted probabilities.
1. Sensitivity analysis: repeat crude and adjusted models with `died` or
   recoded `x60d_mortality`; compare direction and rough magnitude.
1. Non-main sheet: inspect row level, identify a stable patient identifier,
   summarize to one row per patient before joining, and document any exclusions.
1. Bibliography: add a `bibliography:` field in YAML and cite at least one
   source in the Introduction or Discussion.
1. Bootstrap: resample rows with replacement, refit the same model many times,
   collect the coefficient of interest, and summarize percentile intervals.
