---
title: "Buffy, Vampires, and Ordinary Differential Equations: Solutions"
subtitle: "Instructor solution set for the guided R and deSolve exercise"
author: "SISMID 2026"
format:
  html:
    toc: true
    code-copy: true
    code-fold: false
  pdf:
    toc: true
    number-sections: true
    geometry:
      - margin=1in
execute:
  echo: true
  message: false
  warning: true
editor: source
---

# Purpose

This is a set of potential solutions for the Buffy ODEs take-home activity.
If you have any questions or issues, please feel free to reach out to Zane
(email wz.billings@gmail.com). These solutions were generated with AI assistance
so I can't guarantee there are no issues here.

::: {.callout-note}
## About this solution set

This document gives one complete set of solutions. Most programming tasks can
be solved in several equivalent ways; the code here emphasizes the same named
vectors, explicit derivative function, base-R graphics, and sanity checks used
in the course ODE module. Numerical answers may differ slightly across solver
versions or tolerance settings.
:::

# Background and model

Sunnydale has two populations:

- $H(t)$: the number of humans at time $t$;
- $V(t)$: the number of vampires at time $t$.

Time is measured in **years**. Treat $H(t)$ and $V(t)$ as continuous expected
population sizes. A model value such as 15.8 vampires means an expected or
long-run average population near 16, not a literal fraction of a vampire.

We will make the following assumptions.

1. Without vampires, the human population grows logistically at an intrinsic
   rate of 0.10 per year and has a carrying capacity of 90,000.
2. Human-vampire encounters occur at rate $aH(t)V(t)$, where
   $a=1/300$.
3. Each encounter removes one human.
4. A proportion $p=1/240$ of those encounters turns the human into a new
   vampire.
5. Buffy and her allies destroy vampires at rate $d=2/3$ per vampire-year.
6. The Hellmouth attracts or creates vampires at rate $m=0.075$ per
   vampire-year.
7. Initially, Sunnydale has 38,500 humans and 35 vampires.

The model is

$$
\begin{aligned}
\frac{dH}{dt}
&=
 rH\left(1-\frac{H}{K}\right)-aHV,\\[4pt]
\frac{dV}{dt}
&=
 paHV-dV+mV.
\end{aligned}
$$

The parameter values are summarized below.

| Symbol | R name | Value | Interpretation |
|---|---|---:|---|
| $r$ | `human_growth_rate` | 0.10 | Intrinsic human growth rate per year |
| $K$ | `carrying_capacity` | 90,000 | Human carrying capacity |
| $a$ | `encounter_rate` | $1/300$ | Human-vampire encounter coefficient |
| $p$ | `turning_probability` | $1/240$ | Proportion of encounters producing a new vampire |
| $d$ | `slaying_rate` | $2/3$ | Vampire removal rate per year |
| $m$ | `hellmouth_rate` | 0.075 | Background vampire recruitment rate per year |

# Task 0: Set up R

The course setup includes `deSolve`. If it is not installed, run the following
once in the R console:

```r
install.packages("deSolve")
```

Load it in your document.

```{r}
#| label: setup

library(deSolve)
```

The [Module 8 ODE notes](../modules/Module08-ODEs.qmd) use the same basic
`deSolve` workflow as this exercise.

# Task 1: Read the equations as flows

Before writing code, identify the biological or narrative meaning of each term.
Complete the table in prose.

| Term | Which population changes? | Does the term add or remove individuals? | Interpretation |
|---|---|---|---|
| $rH(1-H/K)$ | Humans | Adds humans when $0<H<K$ | Logistic human population growth, limited by carrying capacity |
| $aHV$ in the human equation | Humans | Removes humans | Humans lost through human-vampire encounters |
| $paHV$ | Vampires | Adds vampires | The fraction $p$ of encounters that turns a human into a vampire |
| $dV$ | Vampires | Removes vampires | Vampires destroyed by Buffy and her allies |
| $mV$ | Vampires | Adds vampires | Background recruitment, immigration, or creation through the Hellmouth |

**Student prompt**

In two or three sentences, explain why the human encounter term is $aHV$ but
the vampire turning term is $paHV$. Why do these terms have different
coefficients even though they arise from the same encounters?

**Hint**

Every encounter removes a human, but only some encounters produce a new
vampire.

**Deliverable**

A completed table and a short explanation of the difference between removal
of humans and production of vampires.

::: {.callout-important}
## Solution

The encounter rate is $aHV$ because encounters become more common when either
population is larger, and every encounter removes one human in this model. Only
a proportion $p$ of encounters creates a new vampire, so the vampire inflow is
$p(aHV)=paHV$. Thus, the same encounter process contributes a full loss to the
human equation but only a fractional gain to the vampire equation.
:::

# Task 2: Create parameters, initial conditions, and output times

`deSolve::ode()` needs:

1. a named vector of initial state values;
2. a vector of times where output should be returned;
3. a derivative function;
4. a named vector of parameters.

The names are important. They let us retrieve values safely inside the model
function and become column names in the result.

The following setup is complete. Run it and inspect each object.

```{r}
#| label: model-inputs

buffy_parameters <- c(
  human_growth_rate = 0.10,
  carrying_capacity = 90000,
  encounter_rate = 1 / 300,
  turning_probability = 1 / 240,
  slaying_rate = 2 / 3,
  hellmouth_rate = 0.075
)

buffy_initial_state <- c(
  humans = 38500,
  vampires = 35
)

buffy_times <- seq(
  from = 0,
  to = 100,
  by = 0.1
)

buffy_parameters
buffy_initial_state
head(buffy_times)
tail(buffy_times)
```

**Student prompt**

Answer the following questions.

1. How many output rows should the final solution contain?
2. What does `by = 0.1` mean in years? Approximately how many days is this?
3. Does `by = 0.1` force the numerical solver to use steps of exactly 0.1
   years internally?

**Hint**

The `times` vector specifies where you want `deSolve` to report results.
Adaptive solvers may take additional internal steps between those times.

**Deliverable**

Three brief answers. Use R to calculate the expected number of rows and the
approximate number of days in 0.1 years.

```{r}
#| label: time-grid-answers

length(buffy_times)
0.1 * 365.25
```

::: {.callout-important}
## Solution

1. The solution contains 1,001 requested output rows, including both year 0
   and year 100.
2. `by = 0.1` requests output every tenth of a year, or about 36.5 days.
3. No. The output grid is not necessarily the solver's internal integration
   grid. `lsoda` adaptively chooses internal step sizes and interpolates the
   solution at the requested output times.
:::

# Task 3: Write the derivative function

A `deSolve` derivative function has this structure:

```r
function(time, state, parameters) {
  # Unpack the state and parameters.
  # Calculate all derivatives.
  # Return the derivatives inside a list.
}
```

The order of the arguments matters. Even if your equations do not explicitly
use `time`, the first argument must still be present.

The completed derivative function below writes each process separately so that
its contribution can be checked and interpreted.

```{r}
#| label: buffy-derivative

buffy_ode <- function(time, state, parameters) {
  # State variables
  H <- state[["humans"]]
  V <- state[["vampires"]]

  # Parameters
  r <- parameters[["human_growth_rate"]]
  K <- parameters[["carrying_capacity"]]
  a <- parameters[["encounter_rate"]]
  p <- parameters[["turning_probability"]]
  d <- parameters[["slaying_rate"]]
  m <- parameters[["hellmouth_rate"]]

  # Individual flows
  human_logistic_growth <- r * H * (1 - H / K)
  human_encounter_loss <- a * H * V
  vampire_turning <- p * human_encounter_loss
  vampire_slaying <- d * V
  vampire_hellmouth_recruitment <- m * V

  # Net rates of change: inflows minus outflows.
  d_humans <- human_logistic_growth - human_encounter_loss
  d_vampires <- vampire_turning - vampire_slaying +
    vampire_hellmouth_recruitment

  list(
    c(
      humans = d_humans,
      vampires = d_vampires
    ),
    human_logistic_growth = human_logistic_growth,
    human_encounter_loss = human_encounter_loss,
    vampire_turning = vampire_turning,
    vampire_slaying = vampire_slaying,
    vampire_hellmouth_recruitment = vampire_hellmouth_recruitment
  )
}
```

::: {.callout-tip}
## Hint 1

A derivative is the sum of all flows into a population minus all flows out of
that population.
:::

::: {.callout-tip}
## Hint 2

For humans, logistic growth is an inflow and vampire encounters are an
outflow. For vampires, turning and Hellmouth recruitment are inflows, while
slaying is an outflow.
:::

::: {.callout-warning}
Do not return only the two numbers. `deSolve` requires the derivatives to be
inside a list: `list(c(humans = ..., vampires = ...))`. Additional named
quantities may follow the derivative vector, as in the template above.
:::

**Deliverable**

A derivative function with valid expressions for `d_humans` and
`d_vampires`.

::: {.callout-important}
## Solution

The completed derivative statements are

```r
d_humans <- human_logistic_growth - human_encounter_loss
d_vampires <- vampire_turning - vampire_slaying +
  vampire_hellmouth_recruitment
```

Equivalent code may substitute the equations directly rather than first
creating the named flow objects.
:::

# Task 4: Check the initial rates before solving

It is good practice to test a derivative function at the initial state before
asking a numerical solver to use it repeatedly.

Evaluate the completed derivative function at the initial state and inspect
the output.

```{r}
#| label: initial-rates

initial_rates <- buffy_ode(
  time = 0,
  state = buffy_initial_state,
  parameters = buffy_parameters
)

initial_rates[[1]]
initial_rates[-1]
```

The first list element contains the derivatives. The remaining elements are
auxiliary outputs describing individual flows.

**Student prompt**

1. Is the human population initially increasing or decreasing?
2. Is the vampire population initially increasing or decreasing?
3. Which two human flows are being balanced?
4. Which three vampire flows are being balanced?
5. Explain the units of `d_humans` and `d_vampires`.

**Hint**

A positive derivative means the population is increasing at that instant. A
negative derivative means it is decreasing. The derivative is a rate, not the
number that will necessarily be added or removed over a full year.

**Deliverable**

The initial derivatives and a short plain-language interpretation of their
signs and units.

::: {.callout-important}
## Solution

At time zero, the component flows are approximately:

- human logistic growth: 2,203.06 humans per year;
- human encounter loss: 4,491.67 humans per year;
- vampire turning: 18.72 vampires per year;
- vampire slaying: 23.33 vampires per year;
- Hellmouth recruitment: 2.63 vampires per year.

Therefore,

$$
\left.\frac{dH}{dt}\right|_{t=0}\approx-2288.61
\qquad\text{and}\qquad
\left.\frac{dV}{dt}\right|_{t=0}\approx-1.993.
$$

Both populations are initially decreasing. The derivative units are humans
per year and vampires per year. They are instantaneous rates at $t=0$, not
one-year population changes.
:::

# Task 5: Solve the initial-value problem with `ode()`

After your derivative function works, solve the model.

```{r}
#| label: solve-model

buffy_solution <- deSolve::ode(
  y = buffy_initial_state,
  times = buffy_times,
  func = buffy_ode,
  parms = buffy_parameters,
  method = "lsoda"
)

buffy_df <- as.data.frame(buffy_solution)

head(buffy_df)
tail(buffy_df)
```

`ode()` returns a matrix-like object. The first column is `time`. State
variables and auxiliary outputs occupy the remaining columns.

**Student prompt**

1. Confirm that the result contains columns named `time`, `humans`, and
   `vampires`.
2. Identify the columns corresponding to the auxiliary flows returned by the
   derivative function.
3. Compare the first row with `buffy_initial_state`. Why should they match?

**Hint**

Use `names(buffy_df)` and `nrow(buffy_df)` to inspect the result.

**Deliverable**

A successfully created `buffy_df` and three brief answers.

::: {.callout-important}
## Solution

The primary columns are `time`, `humans`, and `vampires`. The remaining columns
are the auxiliary flows: `human_logistic_growth`, `human_encounter_loss`,
`vampire_turning`, `vampire_slaying`, and
`vampire_hellmouth_recruitment`. The first row matches the initial state because
an initial-value problem is constrained to equal the supplied state at time
zero.
:::

# Task 6: Perform sanity checks

A numerical result is not automatically a trustworthy result. Check the
solution before interpreting it.

Complete the missing checks below.

```{r}
#| label: sanity-checks

# Check the structure.
nrow(buffy_df)
names(buffy_df)

# Check that all state values are finite.
all(is.finite(buffy_df$humans))
all(is.finite(buffy_df$vampires))

# Check that both populations remain non-negative.
all(buffy_df$humans >= 0)
all(buffy_df$vampires >= 0)

# Examine the simulated ranges.
range(buffy_df$humans)
range(buffy_df$vampires)

# Find the row where the human population reaches its minimum.
human_min_row <- which.min(buffy_df$humans)
buffy_df[human_min_row, c("time", "humans", "vampires")]

# Find and print the row where vampires reach their minimum.
vampire_min_row <- which.min(buffy_df$vampires)
buffy_df[vampire_min_row, c("time", "humans", "vampires")]
```

**Student prompt**

1. Did either simulated population become negative or non-finite?
2. When did the human population reach its minimum?
3. When did the vampire population reach its minimum?
4. Which population reached its minimum first?
5. Does the ordering make sense for a predator-prey-like system? Explain.

**Hint**

Changes in one population affect the other with a delay. Do not expect the two
minima to occur at the same time.

**Deliverable**

Completed checks and a short interpretation of the timing of the two minima.

::: {.callout-important}
## Solution

Neither population becomes negative or non-finite. With output every 0.1
years, humans reach their first and global minimum over the first century near
year 5.5, at about 32,834 humans. Vampires reach their minimum later, near year
16.8, at about 7.48 vampires. The human minimum occurs first: humans decline
while vampire abundance is still substantial, and the resulting scarcity of
humans later causes the vampire population to bottom out. This lag is typical
of predator-prey-like feedback.
:::

# Task 7: Plot both populations over time

Humans and vampires are on very different numerical scales, so separate panels
are clearer than placing both variables on one ordinary y-axis.

Complete the second plot and add useful labels.

```{r}
#| label: time-series-plots
#| fig-width: 8
#| fig-height: 7

old_par <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(
  buffy_df$time,
  buffy_df$humans,
  type = "l",
  lwd = 2,
  xlab = "Time (years)",
  ylab = "Humans",
  main = "Human population"
)

abline(h = 42600, lty = 2)

plot(
  buffy_df$time,
  buffy_df$vampires,
  type = "l",
  lwd = 2,
  xlab = "Time (years)",
  ylab = "Vampires",
  main = "Vampire population"
)
abline(h = 15.8, lty = 2)

par(old_par)
```

**Student prompt**

Describe the trajectories using complete sentences.

- Which population declines first?
- Do the populations appear to oscillate?
- Does the size of the oscillation appear constant, increasing, or decreasing?
- By year 100, does the system appear to have settled completely?

**Hint**

An oscillation may be damped: later peaks and valleys can be smaller than
earlier peaks and valleys.

**Deliverable**

A two-panel time-series figure and a paragraph interpreting the main pattern.

::: {.callout-important}
## Solution

Humans decline first because the initial vampire population creates encounter
losses faster than logistic growth replaces people. Vampires also initially
decline, but their later trajectory responds to the changing human population.
Both populations oscillate, with later peaks and valleys smaller than earlier
ones. This is a damped oscillation. At year 100 the state is approximately
$(43{,}741,16.01)$, which is close to but not exactly at equilibrium.
:::

# Task 8: Create a phase-plane plot

A phase-plane plot removes time from the axes and plots one state variable
against another. Each point represents the state of the whole system at one
time.

```{r}
#| label: phase-plane
#| fig-width: 7
#| fig-height: 6

plot(
  buffy_df$humans,
  buffy_df$vampires,
  type = "l",
  lwd = 2,
  xlab = "Humans",
  ylab = "Vampires",
  main = "Sunnydale phase plane"
)

# Mark the starting point.
points(
  buffy_df$humans[1],
  buffy_df$vampires[1],
  pch = 16,
  cex = 1.2
)

# Mark the final simulated point with a different plotting symbol.
points(
  tail(buffy_df$humans, 1),
  tail(buffy_df$vampires, 1),
  pch = 17,
  cex = 1.2
)

# Mark the coexistence equilibrium.
points(42600, 15.8, pch = 4, cex = 1.5, lwd = 2)

# Optional: add an arrow showing the direction near the beginning.
# Choose two nearby rows, such as rows 1 and 15.
arrows(
  x0 = buffy_df$humans[1],
  y0 = buffy_df$vampires[1],
  x1 = buffy_df$humans[15],
  y1 = buffy_df$vampires[15],
  length = 0.10
)
```

**Student prompt**

1. Does the trajectory form a closed loop, move outward, or spiral inward?
2. What would each of those three patterns imply about long-run behavior?
3. How does the phase-plane plot support or revise your interpretation from
   the time-series plots?

**Hint**

A trajectory that spirals inward suggests damped oscillation toward a steady
state. A closed loop suggests persistent oscillation. A trajectory moving
outward suggests instability.

**Deliverable**

A labeled phase-plane plot and a short interpretation.

::: {.callout-important}
## Solution

The trajectory spirals inward. A closed loop would indicate persistent
oscillation of constant amplitude, an outward spiral would indicate an
unstable equilibrium, and an inward spiral indicates damped oscillation toward
an attracting equilibrium. The phase plane therefore confirms the impression
from the time-series plots that the system approaches a stable coexistence
state rather than oscillating forever with constant amplitude.
:::

# Task 9: Find the steady states

A steady state $(H^*,V^*)$ satisfies

$$
\frac{dH}{dt}=0
\qquad\text{and}\qquad
\frac{dV}{dt}=0.
$$

The equations can be factored as

$$
\frac{dH}{dt}
=
H\left[r\left(1-\frac{H}{K}\right)-aV\right]
$$

and

$$
\frac{dV}{dt}
=
V\left[paH-d+m\right].
$$

This immediately gives two boundary equilibria:

- extinction of both populations: $(0,0)$;
- humans at carrying capacity with no vampires: $(K,0)$.

For a coexistence equilibrium, assume $H^*>0$ and $V^*>0$. Then the expressions
inside both square brackets must equal zero.

The completed R code below uses the factored equations.

```{r}
#| label: coexistence-equilibrium

r <- buffy_parameters[["human_growth_rate"]]
K <- buffy_parameters[["carrying_capacity"]]
a <- buffy_parameters[["encounter_rate"]]
p <- buffy_parameters[["turning_probability"]]
d <- buffy_parameters[["slaying_rate"]]
m <- buffy_parameters[["hellmouth_rate"]]

# From paH* - d + m = 0:
H_star <- (d - m) / (p * a)

# From r(1 - H*/K) - aV* = 0:
V_star <- (r / a) * (1 - H_star / K)

coexistence_equilibrium <- c(
  humans = H_star,
  vampires = V_star
)

coexistence_equilibrium
```

::: {.callout-tip}
## Algebra hint

Rearrange the vampire equation to isolate $H^*$. Then substitute that result
into the human equation and isolate $V^*$.
:::

Check whether the derivatives are approximately zero at your proposed
coexistence equilibrium.

```{r}
#| label: equilibrium-check

equilibrium_rates <- buffy_ode(
  time = 0,
  state = coexistence_equilibrium,
  parameters = buffy_parameters
)

equilibrium_rates[[1]]
```

Because calculations use floating-point arithmetic, values extremely close to
zero are acceptable.

**Student prompt**

1. Report all three nonnegative equilibria.
2. Which equilibrium appears relevant to the simulation beginning at
   $(38{,}500,35)$?
3. Compare the year-100 state with the coexistence equilibrium.
4. Explain how to interpret a noninteger equilibrium vampire count.

**Deliverable**

The three equilibria, a numerical derivative check, and a short interpretation.

::: {.callout-important}
## Solution

The nonnegative equilibria are

$$
(0,0),\qquad(90{,}000,0),\qquad(42{,}600,15.8).
$$

The coexistence equilibrium is relevant to the supplied initial condition.
Evaluating `buffy_ode()` there returns derivatives equal to zero up to floating-
point error. At year 100, the numerical state is about $(43{,}741,16.01)$, so
it is near but not yet exactly at coexistence. The value 15.8 is an expected or
long-run average abundance, not a literal fractional individual.

The boundary equilibrium $(90{,}000,0)$ is unstable to vampire invasion under
the original parameters because a small vampire population has positive
per-capita growth when humans are near carrying capacity.
:::

# Task 10: Add the equilibrium to the figures

Add horizontal reference lines for the coexistence equilibrium to the two
time-series plots, and add the equilibrium point to the phase-plane plot.

Template commands are shown below. Insert them in the appropriate plotting
chunks rather than running them alone.

```{r}
#| label: equilibrium-marked-plots
#| fig-width: 8
#| fig-height: 7

old_par <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(
  buffy_df$time, buffy_df$humans,
  type = "l", lwd = 2,
  xlab = "Time (years)", ylab = "Humans",
  main = "Humans and coexistence value"
)
abline(h = H_star, lty = 2)

plot(
  buffy_df$time, buffy_df$vampires,
  type = "l", lwd = 2,
  xlab = "Time (years)", ylab = "Vampires",
  main = "Vampires and coexistence value"
)
abline(h = V_star, lty = 2)

par(old_par)
```

```{r}
#| label: equilibrium-marked-phase-plane
#| fig-width: 7
#| fig-height: 6

plot(
  buffy_df$humans, buffy_df$vampires,
  type = "l", lwd = 2,
  xlab = "Humans", ylab = "Vampires",
  main = "Phase plane with coexistence equilibrium"
)
points(buffy_df$humans[1], buffy_df$vampires[1], pch = 16)
points(tail(buffy_df$humans, 1), tail(buffy_df$vampires, 1), pch = 17)
points(H_star, V_star, pch = 4, cex = 1.5, lwd = 2)
```

**Student prompt**

Do the numerical trajectories cross the equilibrium values multiple times?
Explain why crossing an equilibrium coordinate is not the same as the full
system reaching equilibrium.

**Hint**

The system is at equilibrium only when **both** state variables simultaneously
equal their equilibrium values and both derivatives are zero.

**Deliverable**

Revised figures with the coexistence equilibrium clearly marked.

::: {.callout-important}
## Solution

Each coordinate crosses its equilibrium reference line multiple times because
the trajectory oscillates around the joint equilibrium. Crossing $H=H^*$ alone
does not imply equilibrium if $V\ne V^*$, and crossing $V=V^*$ alone does not
imply equilibrium if $H\ne H^*$. The full system is at equilibrium only when
both coordinates have their equilibrium values simultaneously and both
derivatives vanish.
:::

# Task 11: Extend the simulation

One hundred years may not be long enough to see the limiting behavior clearly.
Repeat the solution through year 500.

```{r}
#| label: long-simulation

long_times <- seq(0, 500, by = 0.25)

long_solution <- deSolve::ode(
  y = buffy_initial_state,
  times = long_times,
  func = buffy_ode,
  parms = buffy_parameters,
  method = "lsoda"
)

long_df <- as.data.frame(long_solution)

old_par <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(
  long_df$time, long_df$humans,
  type = "l", lwd = 2,
  xlab = "Time (years)", ylab = "Humans",
  main = "Human population through year 500"
)
abline(h = H_star, lty = 2)

plot(
  long_df$time, long_df$vampires,
  type = "l", lwd = 2,
  xlab = "Time (years)", ylab = "Vampires",
  main = "Vampire population through year 500"
)
abline(h = V_star, lty = 2)

par(old_par)

long_final_state <- unlist(
  long_df[nrow(long_df), c("humans", "vampires")]
)

rbind(
  final_simulated_state = long_final_state,
  coexistence_equilibrium = coexistence_equilibrium,
  difference = long_final_state - coexistence_equilibrium
)
```

A useful numerical comparison is the distance between the simulated state and
the equilibrium:

$$
D(t)=\sqrt{[H(t)-H^*]^2+[V(t)-V^*]^2}.
$$

Because humans and vampires have different scales, this raw distance is mostly
determined by the human difference. For a scale-adjusted comparison, use
relative differences:

$$
D_{\mathrm{rel}}(t)
=
\sqrt{
\left[\frac{H(t)-H^*}{H^*}\right]^2+
\left[\frac{V(t)-V^*}{V^*}\right]^2
}.
$$

The completed calculation is shown below.

```{r}
#| label: equilibrium-distance

long_df$relative_distance <- sqrt(
  ((long_df$humans - H_star) / H_star)^2 +
    ((long_df$vampires - V_star) / V_star)^2
)

plot(
  long_df$time,
  long_df$relative_distance,
  type = "l",
  log = "y",
  xlab = "Time (years)",
  ylab = "Relative distance from equilibrium",
  main = "Approach to the coexistence equilibrium"
)
```

**Student prompt**

1. Does the distance from equilibrium generally decline?
2. Why does it not necessarily decline smoothly at every time point?
3. Based on the simulation, would you describe the coexistence equilibrium as
   attracting or repelling nearby trajectories?

**Deliverable**

A long-run trajectory plot, a distance-to-equilibrium plot, and a short
stability interpretation based on simulation.

::: {.callout-important}
## Solution

The relative distance generally declines, and by year 500 the simulated state
is extremely close to $(42{,}600,15.8)$. It does not decline monotonically at
every output time because the trajectory spirals: it can move temporarily
farther from the equilibrium in one coordinate while the overall oscillation
shrinks. The long-run simulation supports classifying the coexistence
equilibrium as attracting.

The semilog distance plot is useful because approximately exponential damping
appears roughly linear on a logarithmic y-axis, although oscillation creates
small departures from a perfectly straight decline.
:::

# Task 12: Investigate a stronger slaying intervention

Now treat the slaying rate as an intervention parameter. Write a helper
function that simulates the model for a supplied value of `slaying_rate`.

The completed helper function is shown below.

```{r}
#| label: simulation-helper

simulate_buffy <- function(slaying_rate, times = buffy_times) {
  scenario_parameters <- buffy_parameters

  # Replace the slaying rate for this scenario.
  scenario_parameters[["slaying_rate"]] <- slaying_rate

  scenario_solution <- deSolve::ode(
    y = buffy_initial_state,
    times = times,
    func = buffy_ode,
    parms = scenario_parameters,
    method = "lsoda"
  )

  scenario_df <- as.data.frame(scenario_solution)
  scenario_df$slaying_rate <- slaying_rate
  scenario_df
}
```

Simulate several scenarios.

```{r}
#| label: intervention-scenarios

slaying_scenarios <- c(
  lower = 0.50,
  original = 2 / 3,
  higher = 0.90,
  very_high = 1.35
)

scenario_results <- lapply(
  slaying_scenarios,
  simulate_buffy
)

names(scenario_results) <- names(slaying_scenarios)
```

Plot the vampire trajectory from each scenario on the same axes.

```{r}
#| label: intervention-plot
#| fig-width: 8
#| fig-height: 5

plot(
  scenario_results[[1]]$time,
  scenario_results[[1]]$vampires,
  type = "l",
  lwd = 2,
  ylim = range(vapply(
    scenario_results,
    function(x) range(x$vampires),
    numeric(2)
  )),
  xlab = "Time (years)",
  ylab = "Vampires",
  main = "Vampire trajectories under different slaying rates"
)

for (i in 2:length(scenario_results)) {
  lines(
    scenario_results[[i]]$time,
    scenario_results[[i]]$vampires,
    lwd = 2,
    lty = i
  )
}

legend(
  "topright",
  legend = paste0(
    names(slaying_scenarios),
    ": d = ",
    round(slaying_scenarios, 3)
  ),
  lty = seq_along(slaying_scenarios),
  lwd = 2,
  bty = "n"
)
```

**Student prompt**

1. How does increasing the slaying rate change the transient oscillations?
2. Which tested values appear to permit long-run coexistence?
3. Which tested values appear to drive vampires toward extinction?
4. Does eliminating vampires necessarily minimize short-run human losses?
   Explain what you see rather than assuming the answer.

**Hint**

The long-run condition for vampire growth can be studied from

$$
\frac{1}{V}\frac{dV}{dt}=paH-d+m.
$$

At the largest feasible human population, substitute $H=K$. Find the value of
$d$ that makes this per-capita growth rate equal to zero.

**Deliverable**

A comparison plot and a paragraph explaining how the intervention changes both
short-run and long-run behavior.

```{r}
#| label: intervention-summary

scenario_summary <- do.call(
  rbind,
  lapply(names(scenario_results), function(scenario_name) {
    x <- scenario_results[[scenario_name]]
    data.frame(
      scenario = scenario_name,
      slaying_rate = x$slaying_rate[1],
      minimum_humans = min(x$humans),
      time_of_minimum_humans = x$time[which.min(x$humans)],
      vampires_at_year_100 = tail(x$vampires, 1),
      humans_at_year_100 = tail(x$humans, 1)
    )
  })
)

scenario_summary
```

::: {.callout-important}
## Solution

Increasing the slaying rate makes the initial vampire decline faster and
usually reduces the depth of the short-run human decline. The scenarios with
$d=0.50$, $2/3$, and $0.90$ all permit positive coexistence because their
implied human coexistence level remains below carrying capacity. The tested
value $d=1.35$ exceeds the invasion threshold and drives vampires toward zero,
while humans approach 90,000.

Eliminating vampires does not prevent all short-run human losses: the initially
present vampires still cause encounters before they disappear. In these tested
scenarios, however, stronger slaying also reduces those losses. The minimum
human populations are approximately 22,352, 32,834, 36,354, and 37,575 as $d$
increases from 0.50 to 1.35.
:::

# Optional challenge: Calculate the invasion threshold

Suppose vampires are nearly absent and humans are near carrying capacity. The
vampire population can invade only if its per-capita growth rate is positive:

$$
paK-d+m>0.
$$

Use R to calculate the critical slaying rate $d_{\mathrm{critical}}$ at which
this expression equals zero.

```{r}
#| label: invasion-threshold

d_critical <- p * a * K + m

d_critical
```

```{r}
#| label: near-threshold-simulations
#| fig-width: 8
#| fig-height: 5

near_threshold_rates <- c(
  below = d_critical - 0.01,
  above = d_critical + 0.01
)

near_threshold_results <- lapply(
  near_threshold_rates,
  simulate_buffy,
  times = seq(0, 300, by = 0.25)
)
names(near_threshold_results) <- names(near_threshold_rates)

plot(
  near_threshold_results$below$time,
  near_threshold_results$below$vampires,
  type = "l", lwd = 2,
  xlab = "Time (years)", ylab = "Vampires",
  main = "Dynamics near the vampire invasion threshold",
  ylim = range(
    near_threshold_results$below$vampires,
    near_threshold_results$above$vampires
  )
)
lines(
  near_threshold_results$above$time,
  near_threshold_results$above$vampires,
  lwd = 2, lty = 2
)
legend(
  "topright",
  legend = c("Below threshold", "Above threshold"),
  lty = c(1, 2), lwd = 2, bty = "n"
)
```

Then simulate values slightly below and slightly above this threshold. Because
an ODE model treats $d$ as an instantaneous rate, a value greater than 1 per
year is mathematically valid. It does **not** mean that more than 100% of the
vampires are removed in one discrete annual event.

**Optional deliverable**

The critical value, two nearby simulations, and a one-paragraph explanation of
why continuous-time rates are not restricted to the interval from 0 to 1.

::: {.callout-important}
## Solution

The threshold is

$$
d_{\mathrm{critical}}=paK+m
=\frac{1}{240}\frac{1}{300}(90{,}000)+0.075
=1.325\ \text{per year}.
$$

Below this value, vampires introduced into a nearly vampire-free Sunnydale can
grow when humans are near carrying capacity. Above it, their per-capita growth
rate is negative even at the largest feasible human population, so extinction
is attracting. A continuous-time hazard or rate can exceed 1 per year because
it is not a one-time proportion. For example, $d=1.35$ corresponds to an
average lifetime from slaying alone of roughly $1/1.35=0.74$ years under a
constant-hazard interpretation.
:::

# Reflection questions

Answer these after completing the simulation.

1. Which model assumption seems least realistic or most consequential?
2. What additional state variable would most improve the model?
3. Which parameter would be hardest to estimate from real data?
4. What observations would you need to distinguish vampire recruitment through
   turning from vampire recruitment through the Hellmouth?
5. Give one epidemiologic system that could be studied with a structurally
   similar two-population ODE model.

::: {.callout-important}
## Potential reflection answers

1. **Least realistic or most consequential assumption:** The bilinear encounter
   term $aHV$ assumes homogeneous random mixing and no saturation in vampire
   hunting. It is likely the most consequential structural assumption.
2. **Useful additional state variable:** A third state for humans who have been
   bitten but are not yet vampires would introduce a biologically meaningful
   delay. Other defensible choices include active Slayers, hidden vampires, or
   age structure.
3. **Hardest parameter to estimate:** The Hellmouth recruitment rate $m$ would
   be difficult to distinguish from undocumented turning or migration.
4. **Required observations:** We would need longitudinal counts or marked-event
   data distinguishing newly turned local humans from vampires arriving or
   appearing through other mechanisms.
5. **Epidemiologic analogue:** A vector-host model can have a structurally
   similar two-population interaction, although realistic infection models
   usually divide each population into susceptible and infected compartments.

These are examples rather than uniquely correct responses. Well-defended
alternatives are acceptable.
:::

# Rendering this document to PDF with TinyTeX

Quarto can render a `.qmd` file to PDF through LaTeX. TinyTeX is a compact TeX
Live distribution and is the recommended option for most Quarto users.

Install TinyTeX once from a terminal:

```bash
quarto install tinytex
```

Check that Quarto can see it:

```bash
quarto list tools
```

From the repository root, render this file to PDF with:

```bash
quarto render exercises/buffy-vampire-odes-desolve.qmd --to pdf
```

You can also use the **Render** menu in RStudio or Positron and choose PDF.
If an older TinyTeX installation can no longer download LaTeX packages, update
it with:

```bash
quarto update tinytex
```

An R-based alternative is:

```r
install.packages("tinytex")
tinytex::install_tinytex()
```

Use either the Quarto installation route or the R route; you do not need to
install TinyTeX twice.

# Solution-set checklist

The completed solution demonstrates each of the following:

- translated each model term into a population flow;
- created named parameter and initial-state vectors;
- completed and tested `buffy_ode()`;
- solved the model with `deSolve::ode()`;
- checked the result for finite and nonnegative state values;
- produced human and vampire time-series plots;
- produced and interpreted a phase-plane plot;
- calculated and checked all nonnegative equilibria;
- extended the simulation to investigate long-run behavior;
- compared multiple slaying-rate interventions;
- rendered your completed Quarto document successfully.
