---
title: "Buffy, Vampires, and Ordinary Differential Equations"
subtitle: "A guided practice exercise with R and deSolve"
author: "SISMID 2026"
format:
  html:
    toc: true
    code-copy: true
    code-fold: false
  pdf:
    toc: true
    number-sections: true
    geometry:
      - margin=1in
execute:
  echo: true
  message: false
  warning: true
editor: source
---

# Purpose

In this exercise, you will build and explore a two-population ordinary
differential equation (ODE) model using `deSolve`. The model tracks humans and
vampires in Sunnydale, the fictional setting of *Buffy the Vampire Slayer*.

The workflow you'll use for this silly example problem is the
same one used for epidemic, ecological, pharmacokinetic, and other dynamical
models.

By the end of the exercise, you should be able to:

1. translate a verbal description into a system of ODEs;
2. write a derivative function in the format required by `deSolve`;
3. solve an initial-value problem with `deSolve::ode()`;
4. inspect and plot a numerical solution;
5. perform basic sanity checks before interpreting a simulation;
6. use simulation to investigate equilibria and interventions.

::: {.callout-note}
## How to use this worksheet

Work through the tasks in order. Most code chunks contain usable scaffolding,
but some lines are marked `TODO`. Chunks that depend on unfinished code use
`#| eval: false` so that the blank worksheet can render without errors.

After completing a chunk, remove its `#| eval: false` line, run the chunk, and
render the document again. Save often.
:::

# Background and model

Buffy the Vampire Slayer follows Buffy Summers, a young woman charged with protecting the fictional town
of Sunnydale from vampires and other supernatural threats. Vampires feed on humans, and some of their
victims are turned into new vampires. Buffy and her allies patrol Sunnydale and destroy as many vampires
as they can, but the town is built over a supernatural gateway called the Hellmouth, which continually
attracts new vampires.

The residents of Sunnydale would like to know whether Buffy’s patrols are enough to keep the vampire
population under contro

Sunnydale has two populations:

- $H(t)$: the number of humans at time $t$;
- $V(t)$: the number of vampires at time $t$.

Time is measured in **years**. Treat $H(t)$ and $V(t)$ as continuous expected
population sizes. A model value such as 15.8 vampires means an expected or
long-run average population near 16, not a literal fraction of a vampire.

We will make the following assumptions.

1. Without vampires, the human population grows logistically at an intrinsic
   rate of 0.10 per year and has a carrying capacity of 90,000.
2. Human-vampire encounters occur at rate $aH(t)V(t)$, where
   $a=1/300$.
3. Each encounter removes one human.
4. A proportion $p=1/240$ of those encounters turns the human into a new
   vampire.
5. Buffy and her allies destroy vampires at rate $d=2/3$ per vampire-year.
6. The Hellmouth attracts or creates vampires at rate $m=0.075$ per
   vampire-year.
7. Initially, Sunnydale has 38,500 humans and 35 vampires.

The model is

$$
\begin{aligned}
\frac{dH}{dt}
&=
 rH\left(1-\frac{H}{K}\right)-aHV,\\[4pt]
\frac{dV}{dt}
&=
 paHV-dV+mV.
\end{aligned}
$$

The parameter values are summarized below.

| Symbol | R name | Value | Interpretation |
|---|---|---:|---|
| $r$ | `human_growth_rate` | 0.10 | Intrinsic human growth rate per year |
| $K$ | `carrying_capacity` | 90,000 | Human carrying capacity |
| $a$ | `encounter_rate` | $1/300$ | Human-vampire encounter coefficient |
| $p$ | `turning_probability` | $1/240$ | Proportion of encounters producing a new vampire |
| $d$ | `slaying_rate` | $2/3$ | Vampire removal rate per year |
| $m$ | `hellmouth_rate` | 0.075 | Background vampire recruitment rate per year |

# Task 0: Set up R

The course setup includes `deSolve`. If it is not installed, run the following
once in the R console:

```r
install.packages("deSolve")
```

Load it in your document.

```{r}
#| label: setup

library(deSolve)
```

The [Module 8 ODE notes](../modules/Module08-ODEs.qmd) use the same basic
`deSolve` workflow as this exercise.

# Task 1: Read the equations as flows

Before writing code, identify the biological or narrative meaning of each term.
Complete the table in prose.

| Term | Which population changes? | Does the term add or remove individuals? | Interpretation |
|---|---|---|---|
| $rH(1-H/K)$ | | | |
| $aHV$ in the human equation | | | |
| $paHV$ | | | |
| $dV$ | | | |
| $mV$ | | | |

**Student prompt**

In two or three sentences, explain why the human encounter term is $aHV$ but
the vampire turning term is $paHV$. Why do these terms have different
coefficients even though they arise from the same encounters?

**Hint**

Every encounter removes a human, but only some encounters produce a new
vampire.

**Deliverable**

A completed table and a short explanation of the difference between removal
of humans and production of vampires.

# Task 2: Create parameters, initial conditions, and output times

`deSolve::ode()` needs:

1. a named vector of initial state values;
2. a vector of times where output should be returned;
3. a derivative function;
4. a named vector of parameters.

The names are important. They let us retrieve values safely inside the model
function and become column names in the result.

The following setup is complete. Run it and inspect each object.

```{r}
#| label: model-inputs

buffy_parameters <- c(
  human_growth_rate = 0.10,
  carrying_capacity = 90000,
  encounter_rate = 1 / 300,
  turning_probability = 1 / 240,
  slaying_rate = 2 / 3,
  hellmouth_rate = 0.075
)

buffy_initial_state <- c(
  humans = 38500,
  vampires = 35
)

buffy_times <- seq(
  from = 0,
  to = 100,
  by = 0.1
)

buffy_parameters
buffy_initial_state
head(buffy_times)
tail(buffy_times)
```

**Student prompt**

Answer the following questions.

1. How many output rows should the final solution contain?
2. What does `by = 0.1` mean in years? Approximately how many days is this?
3. Does `by = 0.1` force the numerical solver to use steps of exactly 0.1
   years internally?

**Hint**

The `times` vector specifies where you want `deSolve` to report results.
Adaptive solvers may take additional internal steps between those times.

**Deliverable**

Three brief answers. Use R to calculate the expected number of rows and the
approximate number of days in 0.1 years.

# Task 3: Write the derivative function

A `deSolve` derivative function has this structure:

```r
function(time, state, parameters) {
  # Unpack the state and parameters.
  # Calculate all derivatives.
  # Return the derivatives inside a list.
}
```

The order of the arguments matters. Even if your equations do not explicitly
use `time`, the first argument must still be present.

Complete the two derivative lines marked `TODO` below. The helper quantities
are written separately so that each process can be checked and interpreted.

```{r}
#| label: buffy-derivative

buffy_ode <- function(time, state, parameters) {
  # State variables
  H <- state[["humans"]]
  V <- state[["vampires"]]

  # Parameters
  r <- parameters[["human_growth_rate"]]
  K <- parameters[["carrying_capacity"]]
  a <- parameters[["encounter_rate"]]
  p <- parameters[["turning_probability"]]
  d <- parameters[["slaying_rate"]]
  m <- parameters[["hellmouth_rate"]]

  # Individual flows
  human_logistic_growth <- r * H * (1 - H / K)
  human_encounter_loss <- a * H * V
  vampire_turning <- p * human_encounter_loss
  vampire_slaying <- d * V
  vampire_hellmouth_recruitment <- m * V

  # TODO: replace NA_real_ with the correct net rates of change.
  d_humans <- NA_real_
  d_vampires <- NA_real_

  list(
    c(
      humans = d_humans,
      vampires = d_vampires
    ),
    human_logistic_growth = human_logistic_growth,
    human_encounter_loss = human_encounter_loss,
    vampire_turning = vampire_turning,
    vampire_slaying = vampire_slaying,
    vampire_hellmouth_recruitment = vampire_hellmouth_recruitment
  )
}
```

::: {.callout-tip}
## Hint 1

A derivative is the sum of all flows into a population minus all flows out of
that population.
:::

::: {.callout-tip}
## Hint 2

For humans, logistic growth is an inflow and vampire encounters are an
outflow. For vampires, turning and Hellmouth recruitment are inflows, while
slaying is an outflow.
:::

::: {.callout-warning}
Do not return only the two numbers. `deSolve` requires the derivatives to be
inside a list: `list(c(humans = ..., vampires = ...))`. Additional named
quantities may follow the derivative vector, as in the template above.
:::

**Deliverable**

A derivative function with valid expressions for `d_humans` and
`d_vampires`.

# Task 4: Check the initial rates before solving

It is good practice to test a derivative function at the initial state before
asking a numerical solver to use it repeatedly.

Complete Task 3 first. Then remove `#| eval: false`, run this chunk, and inspect
the output.

```{r}
#| label: initial-rates
#| eval: false

initial_rates <- buffy_ode(
  time = 0,
  state = buffy_initial_state,
  parameters = buffy_parameters
)

initial_rates[[1]]
initial_rates[-1]
```

The first list element contains the derivatives. The remaining elements are
auxiliary outputs describing individual flows.

**Student prompt**

1. Is the human population initially increasing or decreasing?
2. Is the vampire population initially increasing or decreasing?
3. Which two human flows are being balanced?
4. Which three vampire flows are being balanced?
5. Explain the units of `d_humans` and `d_vampires`.

**Hint**

A positive derivative means the population is increasing at that instant. A
negative derivative means it is decreasing. The derivative is a rate, not the
number that will necessarily be added or removed over a full year.

**Deliverable**

The initial derivatives and a short plain-language interpretation of their
signs and units.

# Task 5: Solve the initial-value problem with `ode()`

After your derivative function works, solve the model.

```{r}
#| label: solve-model
#| eval: false

buffy_solution <- deSolve::ode(
  y = buffy_initial_state,
  times = buffy_times,
  func = buffy_ode,
  parms = buffy_parameters,
  method = "lsoda"
)

buffy_df <- as.data.frame(buffy_solution)

head(buffy_df)
tail(buffy_df)
```

`ode()` returns a matrix-like object. The first column is `time`. State
variables and auxiliary outputs occupy the remaining columns.

**Student prompt**

1. Confirm that the result contains columns named `time`, `humans`, and
   `vampires`.
2. Identify the columns corresponding to the auxiliary flows returned by the
   derivative function.
3. Compare the first row with `buffy_initial_state`. Why should they match?

**Hint**

Use `names(buffy_df)` and `nrow(buffy_df)` to inspect the result.

**Deliverable**

A successfully created `buffy_df` and three brief answers.

# Task 6: Perform sanity checks

A numerical result is not automatically a trustworthy result. Check the
solution before interpreting it.

Complete the missing checks below.

```{r}
#| label: sanity-checks
#| eval: false

# Check the structure.
nrow(buffy_df)
names(buffy_df)

# Check that all state values are finite.
all(is.finite(buffy_df$humans))
all(is.finite(buffy_df$vampires))

# TODO: check that both populations remain non-negative.
all(buffy_df$humans >= 0)
# Add the corresponding vampire check here.

# Examine the simulated ranges.
range(buffy_df$humans)
range(buffy_df$vampires)

# Find the row where the human population reaches its minimum.
human_min_row <- which.min(buffy_df$humans)
buffy_df[human_min_row, c("time", "humans", "vampires")]

# TODO: find and print the row where vampires reach their minimum.
```

**Student prompt**

1. Did either simulated population become negative or non-finite?
2. When did the human population reach its minimum?
3. When did the vampire population reach its minimum?
4. Which population reached its minimum first?
5. Does the ordering make sense for a predator-prey-like system? Explain.

**Hint**

Changes in one population affect the other with a delay. Do not expect the two
minima to occur at the same time.

**Deliverable**

Completed checks and a short interpretation of the timing of the two minima.

# Task 7: Plot both populations over time

Humans and vampires are on very different numerical scales, so separate panels
are clearer than placing both variables on one ordinary y-axis.

Complete the second plot and add useful labels.

```{r}
#| label: time-series-plots
#| eval: false
#| fig-width: 8
#| fig-height: 7

old_par <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(
  buffy_df$time,
  buffy_df$humans,
  type = "l",
  lwd = 2,
  xlab = "Time (years)",
  ylab = "Humans",
  main = "Human population"
)

# TODO: make a corresponding line plot for vampires.

par(old_par)
```

**Student prompt**

Describe the trajectories using complete sentences.

- Which population declines first?
- Do the populations appear to oscillate?
- Does the size of the oscillation appear constant, increasing, or decreasing?
- By year 100, does the system appear to have settled completely?

**Hint**

An oscillation may be damped: later peaks and valleys can be smaller than
earlier peaks and valleys.

**Deliverable**

A two-panel time-series figure and a paragraph interpreting the main pattern.

# Task 8: Create a phase-plane plot

A phase-plane plot removes time from the axes and plots one state variable
against another. Each point represents the state of the whole system at one
time.

```{r}
#| label: phase-plane
#| eval: false
#| fig-width: 7
#| fig-height: 6

plot(
  buffy_df$humans,
  buffy_df$vampires,
  type = "l",
  lwd = 2,
  xlab = "Humans",
  ylab = "Vampires",
  main = "Sunnydale phase plane"
)

# Mark the starting point.
points(
  buffy_df$humans[1],
  buffy_df$vampires[1],
  pch = 16,
  cex = 1.2
)

# TODO: mark the final simulated point with a different plotting symbol.

# Optional: add an arrow showing the direction near the beginning.
# Choose two nearby rows, such as rows 1 and 15.
arrows(
  x0 = buffy_df$humans[1],
  y0 = buffy_df$vampires[1],
  x1 = buffy_df$humans[15],
  y1 = buffy_df$vampires[15],
  length = 0.10
)
```

**Student prompt**

1. Does the trajectory form a closed loop, move outward, or spiral inward?
2. What would each of those three patterns imply about long-run behavior?
3. How does the phase-plane plot support or revise your interpretation from
   the time-series plots?

**Hint**

A trajectory that spirals inward suggests damped oscillation toward a steady
state. A closed loop suggests persistent oscillation. A trajectory moving
outward suggests instability.

**Deliverable**

A labeled phase-plane plot and a short interpretation.

# Task 9: Find the steady states

A steady state $(H^*,V^*)$ satisfies

$$
\frac{dH}{dt}=0
\qquad\text{and}\qquad
\frac{dV}{dt}=0.
$$

The equations can be factored as

$$
\frac{dH}{dt}
=
H\left[r\left(1-\frac{H}{K}\right)-aV\right]
$$

and

$$
\frac{dV}{dt}
=
V\left[paH-d+m\right].
$$

This immediately gives two boundary equilibria:

- extinction of both populations: $(0,0)$;
- humans at carrying capacity with no vampires: $(K,0)$.

For a coexistence equilibrium, assume $H^*>0$ and $V^*>0$. Then the expressions
inside both square brackets must equal zero.

Complete the R code below using the factored equations.

```{r}
#| label: coexistence-equilibrium
#| eval: false

r <- buffy_parameters[["human_growth_rate"]]
K <- buffy_parameters[["carrying_capacity"]]
a <- buffy_parameters[["encounter_rate"]]
p <- buffy_parameters[["turning_probability"]]
d <- buffy_parameters[["slaying_rate"]]
m <- buffy_parameters[["hellmouth_rate"]]

# From paH* - d + m = 0:
H_star <- NA_real_  # TODO

# From r(1 - H*/K) - aV* = 0:
V_star <- NA_real_  # TODO

coexistence_equilibrium <- c(
  humans = H_star,
  vampires = V_star
)

coexistence_equilibrium
```

::: {.callout-tip}
## Algebra hint

Rearrange the vampire equation to isolate $H^*$. Then substitute that result
into the human equation and isolate $V^*$.
:::

Check whether the derivatives are approximately zero at your proposed
coexistence equilibrium.

```{r}
#| label: equilibrium-check
#| eval: false

equilibrium_rates <- buffy_ode(
  time = 0,
  state = coexistence_equilibrium,
  parameters = buffy_parameters
)

equilibrium_rates[[1]]
```

Because calculations use floating-point arithmetic, values extremely close to
zero are acceptable.

**Student prompt**

1. Report all three nonnegative equilibria.
2. Which equilibrium appears relevant to the simulation beginning at
   $(38{,}500,35)$?
3. Compare the year-100 state with the coexistence equilibrium.
4. Explain how to interpret a noninteger equilibrium vampire count.

**Deliverable**

The three equilibria, a numerical derivative check, and a short interpretation.

# Task 10: Add the equilibrium to the figures

Add horizontal reference lines for the coexistence equilibrium to the two
time-series plots, and add the equilibrium point to the phase-plane plot.

Template commands are shown below. Insert them in the appropriate plotting
chunks rather than running them alone.

```{r}
#| eval: false

# On the human time-series plot:
abline(h = H_star, lty = 2)

# On the vampire time-series plot:
abline(h = V_star, lty = 2)

# On the phase-plane plot:
points(H_star, V_star, pch = 4, cex = 1.5, lwd = 2)
```

**Student prompt**

Do the numerical trajectories cross the equilibrium values multiple times?
Explain why crossing an equilibrium coordinate is not the same as the full
system reaching equilibrium.

**Hint**

The system is at equilibrium only when **both** state variables simultaneously
equal their equilibrium values and both derivatives are zero.

**Deliverable**

Revised figures with the coexistence equilibrium clearly marked.

# Task 11: Extend the simulation

One hundred years may not be long enough to see the limiting behavior clearly.
Repeat the solution through year 500.

```{r}
#| label: long-simulation
#| eval: false

long_times <- seq(0, 500, by = 0.25)

long_solution <- deSolve::ode(
  y = buffy_initial_state,
  times = long_times,
  func = buffy_ode,
  parms = buffy_parameters,
  method = "lsoda"
)

long_df <- as.data.frame(long_solution)

# TODO: plot humans and vampires through year 500.
# TODO: compare the last row with coexistence_equilibrium.
```

A useful numerical comparison is the distance between the simulated state and
the equilibrium:

$$
D(t)=\sqrt{[H(t)-H^*]^2+[V(t)-V^*]^2}.
$$

Because humans and vampires have different scales, this raw distance is mostly
determined by the human difference. For a scale-adjusted comparison, use
relative differences:

$$
D_{\mathrm{rel}}(t)
=
\sqrt{
\left[\frac{H(t)-H^*}{H^*}\right]^2+
\left[\frac{V(t)-V^*}{V^*}\right]^2
}.
$$

Complete the calculation.

```{r}
#| label: equilibrium-distance
#| eval: false

long_df$relative_distance <- sqrt(
  ((long_df$humans - H_star) / H_star)^2 +
    # TODO: add the squared relative vampire difference.
    0
)

plot(
  long_df$time,
  long_df$relative_distance,
  type = "l",
  log = "y",
  xlab = "Time (years)",
  ylab = "Relative distance from equilibrium",
  main = "Approach to the coexistence equilibrium"
)
```

**Student prompt**

1. Does the distance from equilibrium generally decline?
2. Why does it not necessarily decline smoothly at every time point?
3. Based on the simulation, would you describe the coexistence equilibrium as
   attracting or repelling nearby trajectories?

**Deliverable**

A long-run trajectory plot, a distance-to-equilibrium plot, and a short
stability interpretation based on simulation.

# Task 12: Investigate a stronger slaying intervention

Now treat the slaying rate as an intervention parameter. Write a helper
function that simulates the model for a supplied value of `slaying_rate`.

Complete the missing lines.

```{r}
#| label: simulation-helper
#| eval: false

simulate_buffy <- function(slaying_rate, times = buffy_times) {
  scenario_parameters <- buffy_parameters

  # TODO: replace the slaying rate in scenario_parameters.
  scenario_parameters[["slaying_rate"]] <- NA_real_

  scenario_solution <- deSolve::ode(
    y = buffy_initial_state,
    times = times,
    func = buffy_ode,
    parms = scenario_parameters,
    method = "lsoda"
  )

  scenario_df <- as.data.frame(scenario_solution)
  scenario_df$slaying_rate <- slaying_rate
  scenario_df
}
```

Simulate several scenarios.

```{r}
#| label: intervention-scenarios
#| eval: false

slaying_scenarios <- c(
  lower = 0.50,
  original = 2 / 3,
  higher = 0.90,
  very_high = 1.35
)

scenario_results <- lapply(
  slaying_scenarios,
  simulate_buffy
)

names(scenario_results) <- names(slaying_scenarios)
```

Plot the vampire trajectory from each scenario on the same axes.

```{r}
#| label: intervention-plot
#| eval: false
#| fig-width: 8
#| fig-height: 5

plot(
  scenario_results[[1]]$time,
  scenario_results[[1]]$vampires,
  type = "l",
  lwd = 2,
  ylim = range(vapply(
    scenario_results,
    function(x) range(x$vampires),
    numeric(2)
  )),
  xlab = "Time (years)",
  ylab = "Vampires",
  main = "Vampire trajectories under different slaying rates"
)

# TODO: use a loop or lapply() to add the remaining trajectories with lines().
# TODO: add a legend.
```

**Student prompt**

1. How does increasing the slaying rate change the transient oscillations?
2. Which tested values appear to permit long-run coexistence?
3. Which tested values appear to drive vampires toward extinction?
4. Does eliminating vampires necessarily minimize short-run human losses?
   Explain what you see rather than assuming the answer.

**Hint**

The long-run condition for vampire growth can be studied from

$$
\frac{1}{V}\frac{dV}{dt}=paH-d+m.
$$

At the largest feasible human population, substitute $H=K$. Find the value of
$d$ that makes this per-capita growth rate equal to zero.

**Deliverable**

A comparison plot and a paragraph explaining how the intervention changes both
short-run and long-run behavior.

# Optional challenge: Calculate the invasion threshold

Suppose vampires are nearly absent and humans are near carrying capacity. The
vampire population can invade only if its per-capita growth rate is positive:

$$
paK-d+m>0.
$$

Use R to calculate the critical slaying rate $d_{\mathrm{critical}}$ at which
this expression equals zero.

```{r}
#| label: invasion-threshold
#| eval: false

d_critical <- NA_real_  # TODO

d_critical
```

Then simulate values slightly below and slightly above this threshold. Because
an ODE model treats $d$ as an instantaneous rate, a value greater than 1 per
year is mathematically valid. It does **not** mean that more than 100% of the
vampires are removed in one discrete annual event.

**Optional deliverable**

The critical value, two nearby simulations, and a one-paragraph explanation of
why continuous-time rates are not restricted to the interval from 0 to 1.

# Reflection questions

Answer these after completing the simulation.

1. Which model assumption seems least realistic or most consequential?
2. What additional state variable would most improve the model?
3. Which parameter would be hardest to estimate from real data?
4. What observations would you need to distinguish vampire recruitment through
   turning from vampire recruitment through the Hellmouth?
5. Give one epidemiologic system that could be studied with a structurally
   similar two-population ODE model.

# Rendering this document to PDF with TinyTeX

Quarto can render a `.qmd` file to PDF through LaTeX. TinyTeX is a compact TeX
Live distribution and is the recommended option for most Quarto users.

Install TinyTeX once from a terminal:

```bash
quarto install tinytex
```

Check that Quarto can see it:

```bash
quarto list tools
```

From the repository root, render this file to PDF with:

```bash
quarto render exercises/buffy-vampire-odes-desolve.qmd --to pdf
```

You can also use the **Render** menu in RStudio or Positron and choose PDF.
If an older TinyTeX installation can no longer download LaTeX packages, update
it with:

```bash
quarto update tinytex
```

An R-based alternative is:

```r
install.packages("tinytex")
tinytex::install_tinytex()
```

Use either the Quarto installation route or the R route; you do not need to
install TinyTeX twice.

# Completion checklist

Before considering the exercise complete, confirm that you have:

- translated each model term into a population flow;
- created named parameter and initial-state vectors;
- completed and tested `buffy_ode()`;
- solved the model with `deSolve::ode()`;
- checked the result for finite and nonnegative state values;
- produced human and vampire time-series plots;
- produced and interpreted a phase-plane plot;
- calculated and checked all nonnegative equilibria;
- extended the simulation to investigate long-run behavior;
- compared multiple slaying-rate interventions;
- rendered your completed Quarto document successfully.
