---
title: "Exercise 2: Polish the QCRC Report"
format: html
editor: visual
---

# Purpose

Exercise 2 picks up where Exercise 1 ended. You will turn yesterday's
renderable draft into a polished Quarto deliverable: a short manuscript-style
report with reproducible data loading, a table or figure, a model, clear
interpretation, and a limitations paragraph.

QCRC remains the recommended dataset. If you chose another course dataset in
Exercise 1, you may continue with it, but your report still needs the same
structure and should explain why the alternate dataset supports your question.
See the [course data README](../data-readme.qmd) if you are still deciding.

# Starting point

Begin with your Exercise 1 file, usually `qcrc-mini-manuscript.qmd`.

Save a Day 2 copy as `qcrc-report-final.qmd` before making major changes. Keep
the old draft so you can recover quickly if something breaks.

Use these support files:

- [Starter Quarto template](qcrc-mini-manuscript-starter.qmd)
- [Minimum viable report checklist and rubric](qcrc-report-checklist-rubric.qmd)

# Task 1: Re-render and triage your draft

**Student prompt**

Open your Exercise 1 report and render it once. Make a short to-do list in the
Appendix with the three highest-priority fixes for today.

Your fixes should usually include:

- one reproducibility fix
- one analysis or interpretation fix
- one formatting or reader-experience fix

**Hint**

If the report does not render, make rendering your first fix. Temporarily set
unfinished chunks to `eval: false` only when you are sure they are not needed
for the required deliverable.

**Deliverable**

A Day 2 report file that renders and includes a short triage list.

# Task 2: Make data loading reusable

**Student prompt**

Move your QCRC loading and cleaning code into a small reusable function inside
your report. Then use that function to create the analysis dataset.

Your function should:

- read the `Main_Dataset` sheet by default
- clean names
- convert numeric-looking character variables that you use
- create readable factors for your main binary variables

**Hint**

Functions from Day 1 and functional programming from Day 2 are useful here
because reports often need the same cleaning logic more than once.

```{r}
#| eval: false

read_qcrc_main <- function(path = here::here("data", "QCRC_FINAL_Deidentified.xlsx")) {
  # read, clean, convert, return
}

qcrc <- read_qcrc_main()
```

**Deliverable**

A named data-loading function and a short Methods sentence explaining what it
does.

# Task 3: Add a workbook inventory

**Student prompt**

Use `readxl::excel_sheets()` and `lapply()` to read or inspect the QCRC workbook
sheets. Add a small appendix table with one row per sheet and columns for sheet
name, number of rows, and number of columns.

**Hint**

You do not need to analyze every sheet. The purpose is to practice functional
programming and document what data are available.

```{r}
#| eval: false

data_file <- here::here("data", "QCRC_FINAL_Deidentified.xlsx")
sheet_names <- readxl::excel_sheets(data_file)
sheet_list <- lapply(sheet_names, \(s) readxl::read_xlsx(data_file, sheet = s))
```

**Deliverable**

An appendix table describing the workbook.

# Task 4: Build the analysis table and figure

**Student prompt**

Create at least one polished descriptive table and one figure that support your
research question.

At minimum:

- table: summarize core variables overall and by 30-day mortality status
- figure: show a distribution or relationship relevant to your outcome,
  exposure, or model

**Hint**

Use the simplest table tool that produces a readable result. `knitr::kable()`
is enough. If you already know `gtsummary`, `gt`, or another table package, you
may use it.

**Deliverable**

One table and one figure with captions and one-sentence interpretations.

# Task 5: Fit crude and adjusted models

**Student prompt**

Fit one crude model and one adjusted model for your main question. For the
recommended QCRC report, use logistic regression with 30-day mortality as the
outcome.

Your report should include:

- the model formula written in the Methods section
- crude and adjusted estimates for the main exposure
- confidence intervals or a clear note if you could not compute them
- a plain-language interpretation

**Hint**

Formula syntax and S3 methods are the bridge from the Day 2 module to your
report.

```{r}
#| eval: false

crude_model <- glm(
  x30d_mortality ~ main_exposure,
  data = qcrc,
  family = binomial
)

adjusted_model <- glm(
  x30d_mortality ~ main_exposure + age + bmi,
  data = qcrc,
  family = binomial
)

summary(adjusted_model)
exp(coef(adjusted_model))
exp(confint.default(adjusted_model))
```

**Deliverable**

A compact model-results table and a Results paragraph.

# Task 6: Polish the Quarto deliverable

**Student prompt**

Revise the report so a busy colleague can understand it without reading your
code.

Your final report should have:

- a specific title
- short Introduction, Methods, Results, and Discussion sections
- code chunk labels and useful chunk options
- captions for tables and figures
- an Appendix for checks, exploratory output, or workbook inventory
- a final render to HTML, Word, PDF, or Typst

**Hint**

You do not need a long report. A strong two-to-four page report is better than a
messy long report. Prioritize the question, data, methods, main result, and
limitations.

**Deliverable**

A polished Quarto report that renders without errors.

# Optional advanced tasks

Choose any that are useful for your report.

1. Compare two adjusted models using AIC or analysis of deviance. Explain why
   model selection is not the same as causal variable selection.
1. Use predictions from the adjusted model to report an absolute risk
   difference for two clinically meaningful profiles.
1. Add a sensitivity analysis using `Died`, `x60d_mortality`, or a complete-case
   dataset.
1. Use a non-main QCRC sheet, such as long labs or oxygen delivery, to create an
   additional exploratory variable. Keep the join conservative and document it.
1. Add a bibliography and cite one clinical or epidemiologic reference that
   helps interpret your report.
1. Bootstrap one model estimate and compare it with the model-based confidence
   interval.

# Exercise 2 checklist

Core requirements:

- starts from the Exercise 1 report
- uses QCRC by default, or clearly justifies another course dataset
- includes reusable data-loading or cleaning code
- includes a workbook inventory using `lapply()` or similar repeated logic
- includes a descriptive table and figure
- fits and interprets crude and adjusted models
- uses formula syntax and at least one S3 generic such as `summary()`, `coef()`,
  `confint.default()`, `anova()`, or `predict()`
- renders as a polished Quarto deliverable

Optional stretch goals are listed in the
[rubric/checklist](qcrc-report-checklist-rubric.qmd).
