---
title: "Exercise 1: Start a QCRC Mini-Manuscript"
format: html
editor: visual
---

# Purpose

Today you will start a short manuscript-style report using the same dataset you
will continue using in Exercise 2. The default dataset is the QCRC COVID-19 ICU
workbook because it connects directly to the Quarto, statistics, iteration, and
functions material from Day 1 and the modeling/formula material from Day 2.

Advanced students may choose another course dataset if it is more useful for
their own work. If you switch, keep the same deliverables and explain why the
alternate dataset better supports your question. See the
[course data README](../data-readme.qmd) before switching.

::: {.callout-note}
The realistic goal is not a finished paper today. The goal is a renderable
Quarto draft with a clear question, reproducible data setup, one descriptive
table or figure, one first-pass analysis, and notes about what still needs work.
:::

# Files you will use

- [Starter Quarto template](qcrc-mini-manuscript-starter.qmd)
- [Minimum viable report checklist and rubric](qcrc-report-checklist-rubric.qmd)
- QCRC workbook: `data/QCRC_FINAL_Deidentified.xlsx`

Suggested output file for your own work: `qcrc-mini-manuscript.qmd`.

# Recommended manuscript question

Use this default question unless you have a better one:

> Among patients in the QCRC COVID-19 ICU dataset, what patient characteristics
> and clinical factors are associated with 30-day mortality?

This question is intentionally broad enough for Day 1. In Exercise 2 you will
make it sharper and turn the draft into a polished report.

# Task 1: Create your report draft

**Student prompt**

Open the starter template and save a working copy as
`qcrc-mini-manuscript.qmd`. Render it once before changing anything.

Your draft should keep these manuscript sections:

- Introduction
- Methods
- Results
- Discussion
- Appendix

**Hint**

Rendering early helps you separate Quarto problems from analysis problems. If
the template does not render, fix that before editing the analysis.

**Deliverable**

A renderable Quarto file with the starter structure in place.

# Task 2: Load and inspect the QCRC data

**Student prompt**

In your draft, load the `Main_Dataset` sheet from the QCRC workbook. Clean the
names, convert at least `bmi`, `crp`, or `d_dimer` from character values to
numeric values if you use them, and print a compact data check.

At minimum, your data check should show:

- the number of rows and columns
- the first few cleaned variable names
- a count of missing values for the variables you plan to analyze

**Hint**

Some numeric-looking values in the workbook are stored as text, and missing
values are sometimes written as `"."`.

```{r}
#| eval: false

to_numeric <- function(x) {
  as.numeric(dplyr::na_if(as.character(x), "."))
}

qcrc <- readxl::read_xlsx(
  here::here("data", "QCRC_FINAL_Deidentified.xlsx"),
  sheet = "Main_Dataset"
) |>
  janitor::clean_names() |>
  dplyr::mutate(
    bmi = to_numeric(bmi),
    crp = to_numeric(crp),
    d_dimer = to_numeric(d_dimer)
  )
```

**Deliverable**

A short "Data source and preparation" paragraph plus a data-check chunk in your
Quarto draft.

# Task 3: Define the manuscript variables

**Student prompt**

Write a one-paragraph introduction that states your research question,
outcome, main exposure, and at least two adjustment or stratification variables.

For the default QCRC question, use:

- outcome: `x30d_mortality`
- candidate exposure: `remdesivir_or_placebo`, `intubated`, `bmi`, or `age`
- adjustment variables: choose two from `age`, `bmi`, `female`, `race`,
  `intubated`, `crp`, or `d_dimer`

**Hint**

Do not claim causality unless you can defend the design. For today, language
like "associated with" is safer than "caused by".

**Deliverable**

A short Introduction section with a question and a plain-language hypothesis.

# Task 4: Build one descriptive table or figure

**Student prompt**

Create one table or figure that helps the reader understand the people in the
QCRC dataset before you model anything.

Choose one:

- a table with counts, percentages, means, or medians for your variables
- a histogram or boxplot for a continuous variable such as age, BMI, ICU length
  of stay, CRP, or D dimer
- a bar chart for a binary/categorical variable such as 30-day mortality,
  intubation, sex, race, or treatment group

**Hint**

For a fast table, start with `aggregate()`, `tapply()`, `table()`, or
`dplyr::summarise()`. For a fast figure, start with `ggplot2::ggplot()`.

```{r}
#| eval: false

table(qcrc$x30d_mortality, qcrc$intubated)

aggregate(
  bmi ~ x30d_mortality,
  data = qcrc,
  FUN = \(x) round(mean(x, na.rm = TRUE), 1)
)
```

**Deliverable**

One numbered table or figure with one sentence interpreting it.

# Task 5: Run one first-pass analysis

**Student prompt**

Run one first-pass statistical analysis that connects to your manuscript
question. Choose one:

- a two-sample t-test for a continuous variable by 30-day mortality group
- a chi-square or Fisher exact test for two categorical variables
- a simple logistic regression for 30-day mortality
- a simple linear regression for a continuous outcome such as ICU length of stay

**Hint**

Use formula syntax whenever possible. Put the outcome on the left of `~` and
the predictor on the right.

```{r}
#| eval: false

t.test(bmi ~ x30d_mortality, data = qcrc)

chisq.test(table(qcrc$intubated, qcrc$x30d_mortality))

glm(x30d_mortality ~ age + bmi, data = qcrc, family = binomial)
```

**Deliverable**

One model or test output plus two to three sentences explaining the result in
plain language.

# Task 6: Write the mini-manuscript draft

**Student prompt**

Turn the pieces above into a rough manuscript draft. Each section can be short:

- Introduction: one paragraph
- Methods: one paragraph about the data and analysis
- Results: one paragraph plus your table or figure
- Discussion: one paragraph with the practical interpretation, one limitation,
  and one next step for Exercise 2

**Hint**

A useful fast-moving professional draft often has bracketed notes where you
need later polish. For example: `[Need to check coding of treatment variable]`.
That is better than pretending the draft is finished.

**Deliverable**

A renderable mini-manuscript draft that Exercise 2 can continue.

# Optional advanced tasks

Try these only after the core draft renders.

1. Read all QCRC workbook sheets using `readxl::excel_sheets()` and `lapply()`.
   Add an appendix table with the sheet names and dimensions.
1. Write a small helper function that returns a mean, SD, and missing count for
   any numeric column.
1. Try a sensitivity analysis using `Died` or `x60d_mortality` instead of
   `x30d_mortality`.
1. If you chose a non-QCRC dataset, add a paragraph explaining what the dataset
   can and cannot answer better than QCRC.

# Exercise 1 checklist

Before you leave today, make sure your draft has:

- a clear question and hypothesis
- QCRC loaded and cleaned, or a documented reason for using another dataset
- one table or figure
- one first-pass test or model
- a short interpretation that uses cautious association language
- a renderable Quarto output
- at least one note about what you will improve in Exercise 2

Use the [rubric/checklist](qcrc-report-checklist-rubric.qmd) to decide what is
"good enough" for today.
