---
title: "QCRC Mini-Manuscript Starter"
subtitle: "Replace this subtitle with your research question"
author: "Your name"
date: today
format:
  html:
    toc: true
    code-fold: show
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(here)
library(janitor)
library(knitr)
library(readxl)
```

# Introduction

Replace this paragraph with a short introduction. State your research question,
your outcome, your main exposure, and why a busy public health or clinical
reader should care.

Default question:

> Among patients in the QCRC COVID-19 ICU dataset, what patient characteristics
> and clinical factors are associated with 30-day mortality?

# Methods

## Data source

The analysis used the `Main_Dataset` sheet from the deidentified QCRC COVID-19
ICU workbook included with the course materials.

```{r}
#| label: load-qcrc

to_numeric <- function(x) {
  as.numeric(dplyr::na_if(as.character(x), "."))
}

read_qcrc_main <- function(path = here::here("data", "QCRC_FINAL_Deidentified.xlsx")) {
  readxl::read_xlsx(path, sheet = "Main_Dataset") |>
    janitor::clean_names() |>
    dplyr::mutate(
      bmi = to_numeric(bmi),
      crp = to_numeric(crp),
      d_dimer = to_numeric(d_dimer),
      vent_los = to_numeric(vent_los),
      x30d_mortality = factor(
        x30d_mortality,
        levels = c(0, 1),
        labels = c("No 30-day mortality", "30-day mortality")
      ),
      died = factor(died, levels = c(0, 1), labels = c("Survived", "Died")),
      intubated = factor(intubated, levels = c(0, 1), labels = c("No", "Yes")),
      remdesivir_or_placebo = factor(
        remdesivir_or_placebo,
        levels = c(0, 1),
        labels = c("Placebo", "Remdesivir")
      ),
      female = factor(female),
      race = factor(race)
    )
}

qcrc <- read_qcrc_main()
```

```{r}
#| label: data-check

data.frame(
  rows = nrow(qcrc),
  columns = ncol(qcrc)
) |>
  knitr::kable(caption = "Basic dimensions of the QCRC analysis dataset.")

names(qcrc)[1:10]
```

## Variables

Replace this paragraph with your outcome, exposure, and adjustment variables.
For the default report, the outcome is 30-day mortality. Candidate exposures
include age, BMI, treatment group, intubation, CRP, or D dimer.

```{r}
#| label: missingness-check

analysis_vars <- c(
  "x30d_mortality",
  "age",
  "bmi",
  "female",
  "intubated",
  "remdesivir_or_placebo"
)

missing_counts <- sapply(qcrc[analysis_vars], \(x) sum(is.na(x)))

data.frame(
  variable = names(missing_counts),
  missing = as.integer(missing_counts)
) |>
  knitr::kable(caption = "Missing values for the planned analysis variables.")
```

## Statistical analysis

Replace this paragraph with your analysis plan. Use cautious language such as
"we estimated associations" unless your design supports causal claims.

# Results

## Descriptive results

```{r}
#| label: descriptive-table

qcrc |>
  dplyr::group_by(x30d_mortality) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_age = round(mean(age, na.rm = TRUE), 1),
    mean_bmi = round(mean(bmi, na.rm = TRUE), 1),
    intubated_percent = round(mean(intubated == "Yes", na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Selected QCRC characteristics by 30-day mortality status.")
```

```{r}
#| label: bmi-by-mortality
#| fig-cap: "BMI distribution by 30-day mortality status in the QCRC dataset."

ggplot(qcrc, aes(x = x30d_mortality, y = bmi)) +
  geom_boxplot(na.rm = TRUE) +
  labs(
    x = NULL,
    y = "BMI"
  )
```

## First-pass model

```{r}
#| label: first-pass-model

model_30d <- glm(
  x30d_mortality ~ age + bmi + intubated + remdesivir_or_placebo,
  data = qcrc,
  family = binomial
)

summary(model_30d)
```

```{r}
#| label: first-pass-odds-ratios

or_table <- data.frame(
  term = names(coef(model_30d)),
  odds_ratio = exp(coef(model_30d)),
  exp(confint.default(model_30d)),
  check.names = FALSE
)

names(or_table) <- c("term", "odds_ratio", "conf_low", "conf_high")

or_table |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), \(x) round(x, 2))
  ) |>
  knitr::kable(caption = "Adjusted odds ratios from a first-pass logistic regression.")
```

Replace this paragraph with a plain-language interpretation of the table. State
which estimate matters most for your question, give its direction and rough
magnitude, and say whether the uncertainty changes your confidence.

# Discussion

Replace this paragraph with the main takeaway, one limitation, and one next
step. A good limitation for this starter report might mention missingness,
variable coding, confounding, small subgroup sizes, or the fact that association
does not prove causation.

# Appendix

## Day 2 to-do list

- Recheck variable coding for the main exposure and outcome.
- Decide whether to add or remove adjustment variables.
- Polish captions, prose, and model interpretation.
