---
title: "Exercise 2"
format: docx
editor: visual
execute:
  echo: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)
library(gtsummary)

qcrc <- readxl::read_excel(
	here::here("data/QCRC_FINAL_Deidentified.xlsx"),
	# Any values that are blank or periods become NAs
	# when we read in the data
	na = c("", ".")
)
```

# Introduction

Having COVID and being in the ICU is not good for you.

In our study, we analyze data from Emory University's ICU. We predicted whether patients died or were discharged from the ICU after being hospitalized with COVID-19 using sex, BMI, whether they got remdesivir, whether they got epoprostonol, whether they were intubated, pressor days, C-reactive protein concentration, hemodialysis days and continuous renal replacement therapy (CRRT) days.

# Methods

```{r}
#| label: data cleaning and formatting

# We have some numeric variables that are actually categorical, so we'll
# format them nicely before we do any calculations.
qcrc_clean <-
	qcrc |>
	# transmute() is like a mutate() and a select() at the same time
	transmute(
		died = factor(
			Died,
			levels = c(0, 1),
			labels = c("No", "Yes")
		),
		died_numeric = Died,
		sex = factor(Female) |> relevel(ref = "Female"),
		bmi = BMI,
		remdesivir = factor(
			Remdesivir_or_placebo,
			levels = c(0, 1),
			labels = c("Placebo", "Remdesivir")
		),
		epoprostenol = factor(
			Epoprostenol,
			levels = c(0, 1),
			labels = c("No", "Yes")
		),
		crp = CRP,
		log_ddimer = log(`D dimer`)
	) |>
	# Only keep complete cases in the data
	# (rows that have no NA values)
	tidyr::drop_na()
```

First, we calculated descriptive statistics for the variables of interest in our dataset, stratified by mortality.

# Results

@tbl-demographics shows the summary statistics for the variables we considered, stratified by whether patients died in the ICU or not.

```{r}
#| label: tbl-demographics
#| tbl-cap: "Demographics and summary statistics for our sample, stratified by whether the patient survived to discharge."

gtsummary::tbl_summary(
	qcrc_clean,
	by = died,
	include = -died_numeric
)
```

We fit a simple logistic regression model to get crude odds ratios for each exposure of interest, and then we estimated all of them in a multivariable odds ratio to get adjusted estimates.

@tbl-table2 shows the logistic regression results.

```{r}
#| label: tbl-table2
#| tbl-cap: "This table shows our crude and adjusted odds ratios."
crude_estimates <- gtsummary::tbl_uvregression(
	qcrc_clean,
	y = died_numeric,
	include = -died,
	method = "glm",
	method.args = list(family = "binomial"),
	exponentiate = TRUE
)

adjusted_model <- glm(
	died_numeric ~ sex + bmi + remdesivir + epoprostenol +
		crp + log_ddimer,
	data = qcrc_clean,
	family = "binomial"
)

adjusted_estimate <-
	adjusted_model |>
	gtsummary::tbl_regression(exponentiate = TRUE)

table2 <- gtsummary::tbl_merge(
	list(crude_estimates, adjusted_estimate),
	tab_spanner = c("Crude", "Adjusted")
)

table2
```

# Discussion
