---
title: "Iteration notes"
format: html
editor: visual
---

## Measles example

"measles" is the variable name for our dataset. You can change the variable name to be anything you want. The part after the `<-` is the actual code.

```{r}
# Meas
meas <- readRDS("data/measles_final.Rds")
str(meas)
```

Get the median, first and third quartiles, and range of measles case counts for each country.

```{r}
# First make an empty loop (preallocation)
countries <- unique(meas$iso3c)
res <- vector(mode = "list", length = length(countries))
names(res) <- countries

# Next loop through every country
for (this_country in countries) {
	# Get the data just for each country
	country_data <- subset(meas, iso3c == this_country)
	
	# Now get those summary statistics that we wanted
	country_quantiles <- quantile(
		country_data$measles_cases,
		c(0.25, 0.5, 0.75),
		na.rm = TRUE
	)
	
	country_range <- range(
		country_data$measles_cases,
		na.rm = TRUE
	)
	
	# Now put all of those things together so we can store them
	# neatly
	out <- data.frame(
		"minimum" = country_range[1],
		"Q1" = country_quantiles[1],
		"median" = country_quantiles[2],
		"Q3" = country_quantiles[3],
		"maximum" = country_range[2]
	)
	
	# Assigning outputs to the correct NAME, not by position
	# To do this right, we have to set the names of the empty
	# list before we run the loop.
	# This is called a FOR - EACH construction.
	# This is "idiomatic" to R -- that means it only works
	# in R and doesn't work well in other languages.
	res[[this_country]] <- out
}
```

Now we'll do the same thing with a "FOR" construction, not a "FOR EACH" construction.

```{r}
res2 <- vector(mode = "list", length = length(countries))

# Next loop through every country
for (i in 1:length(countries)) {
	# Get the country code associated with the current index
	this_country <- countries[i]
	
	# Get the data just for each country
	country_data <- subset(meas, iso3c == this_country)
	
	# Now get those summary statistics that we wanted
	country_quantiles <- quantile(
		country_data$measles_cases,
		c(0.25, 0.5, 0.75),
		na.rm = TRUE
	)
	
	country_range <- range(
		country_data$measles_cases,
		na.rm = TRUE
	)
	
	# Now put all of those things together so we can store them
	# neatly
	out <- data.frame(
		"minimum" = country_range[1],
		"Q1" = country_quantiles[1],
		"median" = country_quantiles[2],
		"Q3" = country_quantiles[3],
		"maximum" = country_range[2]
	)
	
	# Assigning outputs to the correct NAME, not by position
	# To do this right, we have to set the names of the empty
	# list before we run the loop.
	# This is called a FOR - EACH construction.
	# This is "idiomatic" to R -- that means it only works
	# in R and doesn't work well in other languages.
	res2[[i]] <- out
}
names(res2) <- countries
```

We can use `do.call()` to combine this into a nice format.

```{r}
country_stats_df <- do.call(rbind, res)
```

```{r}
rbind(
	res[[1]],
	res[[2]]
)
```

What is `do.call()` doing for us?

```{r}
#| eval: false

out <- do.call(rbind, res)
# Is the same thing as 
out <- rbind(res[[1]], res[[2]], ..., res[[length(res)]])
rownames(out) <- names(res)

```

## Next loop problem for the measles dataset

1.  Calculate the incidence per 10k people for each country/year in the dataset.
2.  Make a plot that shows incidence per 10k people vs. year, with a separate line for each country.

$$
\text{incidence per } 10000 \text{ people} = \frac{\text{Cases}}{\text{Population}} \times 10{,}000
$$

```{r}
# Calculate incidence per 10k, this is vectorized!!
meas$incidence_per_10k <-
	meas$measles_cases / meas$population * 10000

# Next, we need to make a blank plot to draw on.
plot(
	x = NULL, y = NULL,
	xlim = range(meas$year),
	ylim = range(meas$incidence_per_10k, na.rm = TRUE),
	xlab = "Year",
	ylab = "Incidence per 10,000"
)

# So now we need to plot a line for each country.
# THis code plots a line for the first country, but you should
# write a loop instead of copying this 213 times.
# country_1_data <- subset(meas, iso3c == countries[[1]])
# lines(
# 	x = country_1_data$year,
# 	y = country_1_data$incidence_per_10k
# )
for (i in 1:length(countries)) {
	this_country <- countries[[i]]
	this_country_data <- subset(meas, iso3c == this_country)
	
	lines(
		x = this_country_data$year,
		y = this_country_data$incidence_per_10k
	)
	Sys.sleep(5)
}

```

Let's run some code manually to see what is happening.

```{r}
plot(
	x = NULL, y = NULL,
	xlim = range(meas$year),
	ylim = range(meas$incidence_per_10k, na.rm = TRUE),
	xlab = "Year",
	ylab = "Incidence per 10,000"
)

# Plot for country 1
country_1 <- countries[[1]]
country_1_data <- subset(meas, iso3c == country_1)

lines(
	x = country_1_data$year,
	y = country_1_data$incidence_per_10k
)

# Plot for country 2
country_2 <- countries[[2]]
country_2_data <- subset(meas, iso3c == country_2)

lines(
	x = country_2_data$year,
	y = country_2_data$incidence_per_10k
)

# Plot for country 3
country_3 <- countries[[3]]
country_3_data <- subset(meas, iso3c == country_3)

lines(
	x = country_3_data$year,
	y = country_3_data$incidence_per_10k
)
```

## What is a function?

-   A function takes an input and gives you an output. It can only give you one output for the same input (usually).

-   In R, functions are little chunks of code that you put something into, it runs that code, and gives you an output. For example, `sum()`.

```{r}
inputs <- c(1, 2, 3)
output <- sum(inputs)
output
```

-   When R sees a function, it looks up the list of code steps it needs to run, and runs them.

```{r}
square <- function(x) {
	return(x ^ 2)
}

square(3)
square(-898)
square(1:10)
```

```{r}
sum <- function(x, y) {
	return(y - x)
}

sum(3, 4)

# this is a consequence of the LGB
# Local
# Global
# Built-in
```

You can also write functions in one line.

```{r}
times_two <- function(x) 2 * x
```

Why returning stuff in functions matters:

```{r}
my_function <- function(x) {
	# Inside of the function is the local scope
	# What happens in the function stays in the function
	x_squared <- x^2
	x_plus_two <- x + 2
	e_to_the_x <- exp(x)
	a_string <- as.character(x)
	
	return(list(x_squared, x_plus_two, e_to_the_x, a_string))
}

my_function(3)
```

# Parts of a function

```{r}
#| eval: false
# The first line of a function looks like this
# function_name <- function(list of arguments)
# and this is called the "Function signature"

# Two ways to write the function "body"
# One line function
# function_name <- function(list of arguments) body

# Multi-line function
# function_name <- function(list of arguments) {
# 	body
# }

# The name of the argument in your function is called
# the "formal argument"
# The value that you put into a formal argument is called
# the "actual argument"

# Formal argument: x
# Actual argument of x: 2
times_two(2)
```

-   Functions can have many different arguments

```{r}
power_law <- function(x, a, b) {
	y <- a * (x ^ b)
	return(y)
}

power_law(2, 2, 3)
```

-   If some actual arguments are very common, we can specify them as the defaults.

```{r}
# The equals signs for the arguments specify the default
# values
# When we specify the argument defaults we HAVE to use one
# equals sign. Not two, and not the arrow.
power_law_with_defaults <- function(x, a = 1, b = 2) {
	y <- a * (x ^ b)
	return(y)
}

power_law_with_defaults(4)
power_law_with_defaults(4, 1, 2)
power_law_with_defaults(4, 2, 3)
```

-   Vectorization gets confusing when you have multiple arguments.

```{r}
# We can vectorize over EVERY argument by default!
power_law_with_defaults(1:3, 2:4, 5:7)
power_law_with_defaults(1, 2, 5)
power_law_with_defaults(2, 3, 6)
```

-   R will RECYCLE values to make sure all vectorized operations get finished.

```{r}
# Good recycling
power_law_with_defaults(
	x= c(100, 2, 7, 9),
	a = 2,
	b = c(3, 4)
)
# R recycles vectors to complete the function call for the
# actual argument that has the most entries
power_law_with_defaults(
	x = c(100, 2, 7, 9),
	a = c(2, 2, 2, 2),
	b = c(3, 4, 3, 4)
)

# Bad recycling
power_law_with_defaults(
	x = c(100, 2, 7, 9),
	a = 2,
	b = c(3, 4, 5)
)
```

# Function example: let's write a function to compute the sum of squared differences

The SSD (sum of squared differences) is defined as

$$
SSD = \sum_{i=1}^n (x_i - \bar{x})^2; \quad i = 1, 2, \ldots, n
$$

```{r}
ssd <- function(x) {
	# Calculate the x_i - x_bar
	x_no_missing <- na.omit(x)
	x_bar <- mean(x_no_missing)
	differences <- x_no_missing - x_bar
	
	# Next square all of the differences
	squared_differences <- differences ^ 2
	
	# And then get the sum of the squared differences
	sum_sq_differences <- sum(squared_differences)
	return(sum_sq_differences)
}

ssd(meas$measles_cases)
```
